---
title: "Ch 1 Larval CADO w/ multiple populations"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = TRUE)
```

### Setup
```{r include=TRUE}
library(ggplot2)
library(tidyr)
library(plyr)
library(dplyr)
library(qvalue)
library(stringr)
library(pcadapt)
library(poolSeq)
library(ACER)
library(ggman)
library(VennDiagram)
library(scales)
library(data.table)
library(patchwork)
```

```{r include=TRUE}
# Function to convert p-values to q-values
Qvalue_convert <- function(table) {


table <- spread(table,GROUP,PVAL)  
  
table$CHR <- table$CHROM

table %>% 
  mutate(CHR = str_replace(CHR, "NC_035780.1", "1")) %>% 
  mutate(CHR = str_replace(CHR, "NC_035781.1", "2")) %>% 
  mutate(CHR = str_replace(CHR, "NC_035782.1", "3")) %>% 
  mutate(CHR = str_replace(CHR, "NC_035783.1", "4")) %>% 
  mutate(CHR = str_replace(CHR, "NC_035784.1", "5")) %>% 
  mutate(CHR = str_replace(CHR, "NC_035785.1", "6")) %>% 
  mutate(CHR = str_replace(CHR, "NC_035786.1", "7")) %>% 
  mutate(CHR = str_replace(CHR, "NC_035787.1", "8")) %>%
  mutate(CHR = str_replace(CHR, "NC_035788.1", "9")) %>% 
  mutate(CHR = str_replace(CHR, "NC_035789.1", "10"))  -> pp1

pp1 <- unite(pp1, "SNP", c("CHROM","BP"),remove = FALSE)

pp1$CHR <- as.numeric(pp1$CHR)  

# Set pi0=1 to equal Benjamini Hochberg
pp1$QSTR <- qvalue(pp1$PSTR, pi0 = 1)$qvalues
pp1$QCON <- qvalue(pp1$PCON, pi0 = 1)$qvalues


return(pp1)
}
```

```{r include=TRUE}
# Function to determine significant loci in various outputs
Significant_subset <- function(pv, alpha, alpha2) {
  # Add new logical columns based on conditions
  pv$Sig.STR <- pv$QSTR < alpha
  pv$Sig.CON <- pv$QCON < alpha
  
  # Subset the dataframe
  ppsig <- subset(pv, QSTR < alpha )
  ppsig <- subset(ppsig, QCON > alpha2)
  
  # Print diagnostics
  print(nrow(pv))
  print(nrow(ppsig))
  print(nrow(ppsig) / nrow(pv))
  
  # Return the modified dataframe
  return(ppsig)
}
```

```{r include=TRUE}
group_and_average <- function(df) {
  df %>%
    dplyr::group_by(SNP) %>%
    dplyr::summarize(
      across(
        c(PCON, PSTR, CHR, QCON, QSTR),
        ~ if (all(is.na(.))) NA else min(., na.rm = TRUE)
      ),
      Sig.STR = any(Sig.STR, na.rm = TRUE),
      CHROM  = first(CHROM),
      BP     = first(BP),
      .groups = "drop"
    )
}
```

```{r include=TRUE}
#cbPaletteSmall <- c("purple", "green", "gold", "red", "orange", "blue", "violet")
cbPaletteSmall <- c("#3F8F7A", "#D3DDF0", "#4169E1", "#5F7BAA", "#F3D6D9", "#8C3A41", "hotpink")

cbPaletteSmall3 <- c("#3F8F7A", "#D8848A", "#5F7BAA")

ARCpalette <- "#3F8F7A"
NEHpalette <- c("#F3D6D9", "#8C3A41", "hotpink")
MVpalette <- c("#D3DDF0", "#4169E1", "#5F7BAA")

treatPalette <- c("#A29BAD", "#4CAF72", "#D29F13")
```

```{r}
sessionInfo()
```


# Visualize across 10,000 random loci
## All populations PCA of random 10k loci
```{bash, eval = FALSE}
mamba activate CADO

cut -f1,2 Ch1.dp25.cov.sync | sort -u  > filtered.loci

mawk '!/CHR/' filtered.loci | shuf -n 50000  | shuf -n 10000 > rand.loci

mawk 'NR==FNR{a[$1,$2]; next} ($1,$2) in a' rand.loci Ch1.dp25.cov.sync | cut --complement -f67- > input.all.rand.sync

```

```{r}
input.all.rand <- read.sync(file="input.all.rand.sync", gen=seq(1,63), repl=seq(1,63), polarization="rising")


input.all.rand.af <- af(input.all.rand,gen=seq(1,63), repl=seq(1,63))

bio.labels <- c(
  "ARC2_CB1","ARC2_CB2","ARC2_CB3","ARC2_SB1","ARC2_SB2","ARC2_SB3","ARC2_T01","ARC2_T02","ARC2_T03",
  "MV1_CB1","MV1_CB2","MV1_CB3","MV1_SB1","MV1_SB2","MV1_SB3","MV1_T01","MV1_T02","MV1_T03",
  "MV2_CB1","MV2_CB2","MV2_CB3","MV2_SB1","MV2_SB2","MV2_SB3","MV2_T01","MV2_T02","MV2_T03",
  "MV3_CB1","MV3_CB2","MV3_CB3","MV3_SB1","MV3_SB2","MV3_SB3","MV3_T01","MV3_T02","MV3_T03",
  "NEH1_CB1","NEH1_CB2","NEH1_CB3","NEH1_SB1","NEH1_SB2","NEH1_SB3","NEH1_T01","NEH1_T02","NEH1_T03",
  "NEH2_CB1","NEH2_CB2","NEH2_CB3","NEH2_SB1","NEH2_SB2","NEH2_SB3","NEH2_T01","NEH2_T02","NEH2_T03",
  "NEH3_CB1","NEH3_CB2","NEH3_CB3","NEH3_SB1","NEH3_SB2","NEH3_SB3","NEH3_T01","NEH3_T02","NEH3_T03"
)
length(bio.labels)  # should be 63

pop.ids <- colnames(input.all.rand.af)  # "F1.R1.freq", ..., "F63.R63.freq"

meta <- data.frame(
  sample_id   = pop.ids,
  bio_label   = bio.labels,
  stringsAsFactors = FALSE
)

# Derive treatment from the biological label
meta$treatment <- ifelse(grepl("_CB", meta$bio_label), "control",
                  ifelse(grepl("_SB", meta$bio_label), "stress", "initial"))
# can also parse by trial  
meta$trial <- sub("_.*", "", meta$bio_label)
# and parse by population
meta$pop <- ifelse(grepl("^ARC", meta$bio_label), "ARC",
                   ifelse(grepl("^MV",  meta$bio_label),  "MV",
                          ifelse(grepl("^NEH", meta$bio_label), "NEH", NA)))
View(meta)

#check that column names in input.all.rand.af and sample_id column in meta table are the same
all.equal(colnames(input.all.rand.af), meta$sample_id)
# should be TRUE

# assign the columns of input.all.rand.af with their proper sample names instead of the sample_id created by inputting the table to poolseq
colnames(input.all.rand.af) <- meta$bio_label
```

shows a strong pattern of population and experiment year (2023 – ARC2, NEH1, MV1, MV2; 2024 – NEH2, NEH3, MV3)
```{r}
# Perform PCA
af_t <- t(input.all.rand.af)

pca_res <- prcomp(af_t, center = TRUE, scale. = TRUE)

# Get scores for plotting
pca_df <- as.data.frame(pca_res$x)
#pca_df$locus <- rownames(input.all.rand.af)  # keep locus IDs if you have them

var_explained <- pca_res$sdev^2 / sum(pca_res$sdev^2) * 100

pca_df$Treatment <- meta$treatment
pca_df$Trial <- meta$trial

pca_rand10k <- ggplot(pca_df, aes(x = PC1, y = PC2, color = Trial, fill=Trial)) +
  geom_point(aes(alpha=1), size =6,shape=21, col="black")+  scale_fill_manual(values=cbPaletteSmall,name="Exp. Trial")+ 
guides(alpha=FALSE) +theme(axis.title.x = element_text(size = 24),axis.title.y = element_text(size = 24)) +
  theme_minimal() +
  labs(
    x = paste0("PC1 (", round(var_explained[1], 1), "%)"),
    y = paste0("PC2 (", round(var_explained[2], 1), "%)"),
    title = "PCA of Allele Frequencies"
  ); pca_rand10k

ggsave(filename = "PCA/ch1_pca_rand10k.png", plot = pca_rand10k, width = 6, height = 5, units = "in", dpi = 300)
```

### PCA of random 10K loci from All Initial Samples
As expected, shows the same strong pattern of population and experiment year seen in the all sample plot
```{r}
#subset data to only T0 initial samples
t0_cols <- grepl("_T0", colnames(input.all.rand.af))
input.T0.rand.af <- input.all.rand.af[, t0_cols]
afi_t <- t(input.T0.rand.af)

#subset metadata table
t0_meta <- subset(meta, meta$treatment == "initial")
View(t0_meta)

# start pca code
pca_i_res <- prcomp(afi_t, center = TRUE, scale. = TRUE)

# Get scores for plotting
pca_i_df <- as.data.frame(pca_i_res$x)

var_explained <- pca_i_res$sdev^2 / sum(pca_i_res$sdev^2) * 100

pca_i_df$Treatment <- t0_meta$treatment
pca_i_df$Trial <- t0_meta$trial

pca_t0rand10k<- ggplot(pca_i_df, aes(x = PC1, y = PC2, color = Trial, fill=Trial)) +
  geom_point(aes(alpha=0.2), size =6,shape=21, col="black")+  scale_fill_manual(values=cbPaletteSmall,name="Exp. Trial")+ 
guides(alpha=FALSE) +theme(axis.title.x = element_text(size = 24),axis.title.y = element_text(size = 24)) +
  theme_classic() +
  labs(
    x = paste0("PC1 (", round(var_explained[1], 1), "%)"),
    y = paste0("PC2 (", round(var_explained[2], 1), "%)"),
    title = "PCA of Allele Frequencies"
  );pca_t0rand10k

ggsave(filename = "PCA/ch1_pca_initial_rand10k.png", plot = pca_t0rand10k, width = 6, height = 5, units = "in", dpi = 300)
```


#### Random loci for each individual experimental population (ARC, NEH, MV)
```{r}
# Ensure meta is keyed by bio_label and in column order
meta <- meta[match(colnames(input.all.rand.af), meta$bio_label), ]
stopifnot(all(meta$bio_label == colnames(input.all.rand.af)))

pops <- c("ARC", "MV", "NEH")

for (pop in pops) {
  # select columns belonging to this population
  cols_idx <- which(meta$pop == pop)
  cat(sprintf("\n== Population %s: found %d sample columns ==\n", pop, length(cols_idx)))
  if (length(cols_idx) < 2) {
    warning(sprintf("Population %s: less than 2 samples found - skipping PCA.", pop))
    next
  }

  sub_mat <- input.all.rand.af[, cols_idx, drop = FALSE]
  cat("  dim(sub_mat) (loci x samples):", dim(sub_mat), "\n")

  # transpose: rows = samples, cols = loci
  mat_t <- t(sub_mat)

  # drop zero-variance loci within this population
  var_cols <- apply(mat_t, 2, var, na.rm = TRUE)
  keep     <- which(var_cols > 0)
  removed_n <- sum(var_cols == 0)
  cat("  loci removed (zero variance):", removed_n, "\n")

  if (length(keep) < 2) {
    warning(sprintf("Population %s: fewer than 2 variable loci after filtering - skipping PCA.", pop))
    next
  }
  mat_t2 <- mat_t[, keep, drop = FALSE]
  cat("  dim(mat_t2) (samples x retained loci):", dim(mat_t2), "\n")

  # PCA
  pca_res <- prcomp(mat_t2, center = TRUE, scale. = TRUE)
  pca_df  <- as.data.frame(pca_res$x)

  # add metadata for these samples
  sample_labels <- rownames(mat_t2)                      # bio_labels
  sample_meta   <- meta[match(sample_labels, meta$bio_label), ]

  pca_df$population <- sample_meta$population            # should all be 'pop'
  pca_df$treatment  <- sample_meta$treatment             # T0 / CB / SB
  pca_df$bio_label  <- sample_meta$bio_label
  pca_df$trial <- sample_meta$trial

  var_explained <- (pca_res$sdev^2) / sum(pca_res$sdev^2) * 100

  # palette for treatment or population; here color by treatment
  if (exists("treatPalette")) {
    fill_scale <- scale_fill_manual(values = treatPalette, name = "Treatment")
  } else {
    fill_scale <- scale_fill_brewer(palette = "Set1", name = "Treatment")
  }

  # plot: within a population, color by treatment
  p <- ggplot(pca_df, aes(x = PC1, y = PC2, fill = treatment)) +
    geom_point(size = 5, shape = 21, color = "black", alpha = 0.8) +
    fill_scale +
    geom_text_repel(aes(label = trial), size = 3) +
    theme_classic() +
    theme(axis.title.x = element_text(size = 14),
          axis.title.y = element_text(size = 14),
          plot.title   = element_text(hjust = 0.5)) +
    labs(
      title = paste0("PCA - Population ", pop),
      x = paste0("PC1 (", round(var_explained[1], 1), "%)"),
      y = paste0("PC2 (", round(var_explained[2], 1), "%)")
    )

  outfile <- paste0("PCA/PCA_population_", pop, ".png")
  ggsave(outfile, plot = p, width = 10, height = 7, dpi = 300)
  cat("  saved:", outfile, "\n")
}

```

#### Random loci for each individual experimental trial (ARC2, NEH1-3, MV1-3)
```{r}
# Ensure meta is keyed by bio_label and in column order
meta <- meta[match(colnames(input.all.rand.af), meta$bio_label), ]
stopifnot(all(meta$bio_label == colnames(input.all.rand.af)))

trials <- c("ARC2", "MV1", "MV2", "MV3", "NEH1", "NEH2", "NEH3")

for (trial in trials) {
  # select columns belonging to this population
  cols_idx <- which(meta$trial == trial)
  cat(sprintf("\n== Population %s: found %d sample columns ==\n", trial, length(cols_idx)))
  if (length(cols_idx) < 2) {
    warning(sprintf("Population %s: less than 2 samples found - skipping PCA.", trial))
    next
  }

  sub_mat <- input.all.rand.af[, cols_idx, drop = FALSE]
  cat("  dim(sub_mat) (loci x samples):", dim(sub_mat), "\n")

  # transpose: rows = samples, cols = loci
  mat_t <- t(sub_mat)

  # drop zero-variance loci within this population
  var_cols <- apply(mat_t, 2, var, na.rm = TRUE)
  keep     <- which(var_cols > 0)
  removed_n <- sum(var_cols == 0)
  cat("  loci removed (zero variance):", removed_n, "\n")

  if (length(keep) < 2) {
    warning(sprintf("Trial %s: fewer than 2 variable loci after filtering - skipping PCA.", trial))
    next
  }
  mat_t2 <- mat_t[, keep, drop = FALSE]
  cat("  dim(mat_t2) (samples x retained loci):", dim(mat_t2), "\n")

  # PCA
  pca_res <- prcomp(mat_t2, center = TRUE, scale. = TRUE)
  pca_df  <- as.data.frame(pca_res$x)

  # add metadata for these samples
  sample_labels <- rownames(mat_t2)                      # bio_labels
  sample_meta   <- meta[match(sample_labels, meta$bio_label), ]

  pca_df$population <- sample_meta$population            # should all be 'pop'
  pca_df$treatment  <- sample_meta$treatment             # T0 / CB / SB
  pca_df$bio_label  <- sample_meta$bio_label
  pca_df$trial <- sample_meta$trial

  var_explained <- (pca_res$sdev^2) / sum(pca_res$sdev^2) * 100

  # palette for treatment or population; here color by treatment
  if (exists("treatPalette")) {
    fill_scale <- scale_fill_manual(values = treatPalette, name = "Treatment")
  } else {
    fill_scale <- scale_fill_brewer(palette = "Set1", name = "Treatment")
  }

  # plot: within a population, color by treatment
  p <- ggplot(pca_df, aes(x = PC1, y = PC2, fill = treatment)) +
    geom_point(size = 5, shape = 21, color = "black", alpha = 0.8) +
    fill_scale +
    theme_classic() +
    theme(axis.title.x = element_text(size = 14),
          axis.title.y = element_text(size = 14),
          plot.title   = element_text(hjust = 0.5)) +
    labs(
      title = paste0("PCA - Exp. Trial ", trial),
      x = paste0("PC1 (", round(var_explained[1], 1), "%)"),
      y = paste0("PC2 (", round(var_explained[2], 1), "%)")
    )

  outfile <- paste0("PCA/PCA_trial_", trial, ".png")
  ggsave(outfile, plot = p, width = 10, height = 7, dpi = 300)
  cat("  saved:", outfile, "\n")
}
```

#### Venn Diagram for called, filtered loci between populations
Haven't updated this code yet just adding in if I want to come back to it later.
Right now it doesn't seem like this will be that informative as most of my variation is driven by trial (7 groups)
```{bash}
# extract the CHROM and POS of snps from each sync file 
cut -f1,2 ARC.dp25.cov.sync > ARC.pos
cut -f1,2 NEH.dp25.cov.sync > NEH.pos
cut -f1,2 MV.dp25.cov.sync > MV.pos

# count overlap between trials 
cat <(echo "SNP") <(cat *.pos | cut -f1,2 | sort | uniq -c | mawk '$1 <2' | mawk '{print $2"_"$3}') > singleton.loci
```

```{r}
# Read and preprocess data
ARC.loci <- read.table("ARC.pos", header = TRUE)
colnames(ARC.loci) <- c("CHROM", "BP")
ARC.loci <- ARC.loci %>% mutate(ARC = TRUE)

NEH.loci <- read.table("NEH.pos", header = TRUE)
colnames(NEH.loci) <- c("CHROM", "BP")
NEH.loci <- NEH.loci %>% mutate(NEH = TRUE)

MV.loci <- read.table("MV.pos", header = TRUE)
colnames(MV.loci) <- c("CHROM", "BP")
MV.loci <- MV.loci %>% mutate(MV = TRUE)

# Full join all data frames to preserve all CHROM and BP combinations
pos_df <- ARC.loci %>%
  full_join(NEH.loci %>% select(CHROM, BP, NEH), by = c("CHROM", "BP")) %>%
  full_join(MV.loci %>% select(CHROM, BP, MV), by = c("CHROM", "BP"))

# Replace NA values in the BLOCK_* columns with FALSE
pos_df <- pos_df %>%
  mutate(
    ARC = ifelse(is.na(ARC), FALSE, ARC),
    NEH = ifelse(is.na(NEH), FALSE, NEH),
    MV = ifelse(is.na(MV), FALSE, MV)
  )


loci.arc <- nrow(subset(pos_df,ARC == TRUE))
loci.neh <- nrow(subset(pos_df,NEH == TRUE))
loci.mv <- nrow(subset(pos_df,MV == TRUE))

loci.arc.neh <- nrow(subset(pos_df,ARC == TRUE & NEH ==TRUE))
loci.arc.mv <- nrow(subset(pos_df,ARC == TRUE & MV ==TRUE))
loci.neh.mv <- nrow(subset(pos_df,NEH == TRUE & MV ==TRUE))
loci.arc.neh.mv <- nrow(subset(pos_df,ARC == TRUE & NEH ==TRUE & MV ==TRUE))

png(filename="Venn_variants.png", type="cairo",units="px", width=4000, height=4000, res=500, bg="transparent")

venn.plot <- draw.triple.venn(
  area1=loci.arc, area2=loci.neh, area3=loci.mv, 
  n12= loci.arc.neh, n13=loci.arc.mv, 
  n23=loci.neh.mv,scaled=TRUE,
  n123=loci.arc.neh.mv, cex=rep(2, 7),
  category = c("ARC", "NEH", "MV"),
  fill = cbPaletteSmall3, cat.cex = rep(2, 3),
  cat.col = cbPaletteSmall3, label.col = rep("white", 7))
dev.off()

```





# Population level outlier testing
## ARC

First normalize with initial sample coverage. The core logic is that the internal standard (IS) pools (the inital samples) act as a reference to normalize and cap coverage, ensuring that outlier detection downstream isn't driven by coverage artifacts rather than true allele frequency signals.
```{bash, eval=FALSE}
source activate CADO

#sum up initial samples
bash ../Scripts/sum.sh <(cut -f1,2,3,10-12 ARC.dp25.cov.sync) > ARC.IS.sync
# makes extra columns with the sum data
paste ARC.IS.sync <(cut -f4 ARC.IS.sync) <(cut -f4 ARC.IS.sync)  > ARC.ISsum3.sync

# look at what col number to put in the compute average step next 
mawk 'NR==1 { for(i=1; i<=NF; i++) print i, $i }' ARC.dp25.cov.sync
mawk -f ../Scripts/add_cov_sync.txt <(cut -f10-12 --complement ARC.dp25.cov.sync) > ARC.sansIS.cov.sync
mawk 'NR==1 { for(i=1; i<=NF; i++) print i, $i }' ARC.sansIS.cov.sync # last column is 15 (mean coverage w/o initial samples)

# compute average coverage on sync file minus initial samples
AV_COV=$(mawk -f ../Scripts/add_cov_sync.txt <(cut -f10-12 --complement ARC.dp25.cov.sync) | mawk '{sum=sum+$15} END {print sum/NR}')
echo $AV_COV

mawk -f ../Scripts/add_cov_sync.txt ARC.ISsum3.sync | mawk -v x=$AV_COV '$7 >= x' | mawk '!/CHR/' | cut -f1-6 > ARC.ISsum3nh.sync

mawk -f ../Scripts/add_cov_sync_IS.txt ARC.ISsum3.sync | mawk -v x=$AV_COV '$7 >= x'> ARC.ISsum3nh.cov.sync

# apply coverage filter back to main sync

#first check col number to apply filter to
paste ARC.dp25.cov.sync <(mawk -f ../Scripts/add_cov_sync.txt ARC.ISsum3.sync ) > ISnormARC.cov.sync
mawk 'NR==1 { for(i=1; i<=NF; i++) print i, $i }' ISnormARC.cov.sync # col 24 (mean cov normalized by initial samples)

# apply initial sample coverage filter
paste ARC.dp25.cov.sync <(mawk -f ../Scripts/add_cov_sync.txt ARC.ISsum3.sync ) | mawk -v x=$AV_COV '$24 >= x' | cut -f1-15 > temp.sync

# ARC samples normalized by initial sample coverage stored in *COV.sync
mv temp.sync ARC.dp25.COV.sync
#cp ARC.dp25.cov.sync ARC.dp25.COV.sync

# cap coverage at IS maxiumum

#remove inital samples and remove old cov stats and recalculates the coverage with only treatment pools
mawk -f ../Scripts/add_cov_sync.txt <(cut -f10-12,13- --complement ARC.dp25.COV.sync) > temp.cov

# pastes pop coverage with minimum inital sample coverage (col 7) to last col (13) if temp.cov and compare treatemtn pool mean cov ($12) to IS min cov ($13)
paste temp.cov <(cut -f7 ARC.ISsum3nh.cov.sync) | mawk -v OFS='\t' '{if ($12 > $13) {$12=$13; print $0} else {print $0}}' | cut -f1-12> ARC.dp25.COV
```
The final output is ARC.dp25.COV, a fully processed, coverage-normalized sync ready for outlier detection.


```{bash, eval=FALSE}
source activate randowm_draw

# subsampling the internal standard 
#python ../scripts/sub_sample.py ARC.dp25.COV ARC.ISsum3nh.sync ARC.ISs.sync
python ../Scripts/sub_sample.py ARC.ISsum3nh.cov.sync ARC.ISsum3nh.sync ARC.ISs.sync

# add heading
cat <(echo -e "CHROM\tPOS\tREF\tIS_RS1\tIS_RS2\tIS_RS3") ARC.ISs.sync > ARC.ISsum.sync

# building treatment specific input files (interleaved with IS columns)
#these IS columns will serve as the null expectation during outlier detection

# Control columns - will need to update columns based on data
paste <(cut -f1,2,3,4 ARC.ISsum.sync) <(cut -f4 ARC.dp25.COV.sync) \
      <(cut -f5 ARC.ISsum.sync) <(cut -f5 ARC.dp25.COV.sync) \
      <(cut -f6 ARC.ISsum.sync) <(cut -f6 ARC.dp25.COV.sync)> ARC.CON.sync

head -1 ARC.CON.sync
tail -n +2 ARC.CON.sync > ARC.CON.input

# Stress columns - will need to update columns based on data
paste <(cut -f1,2,3,4 ARC.ISsum.sync)  <(cut -f7 ARC.dp25.COV.sync) \
      <(cut -f5 ARC.ISsum.sync) <(cut -f8 ARC.dp25.COV.sync) \
      <(cut -f6 ARC.ISsum.sync) <(cut -f9 ARC.dp25.COV.sync)> ARC.STR.sync

head -1 ARC.STR.sync
tail -n +2 ARC.STR.sync > ARC.STR.input
```

### Calculate P-values - ARC
#### Modified CMH - ARC population

Reads in the interleaved IS+treatment sync files into R and runs an adapted Cochran-Mantel-Haenszel (CMH) test to identify loci where allele frequencies shift significantly between initial samples and treatment pools.

Reading and parsing files - read.sync() requires a headerless input file
```{r}
# gen assigns each column a "generation," where 0 = IS (baseline) and 1 = treatment pool
# repl  pairs each IS replicate with its corresponding treatment pool as replicate 1, 2, and 3
# coverage() and af() extract per-locus read depth and allele frequencies from the parsed sync object, which are passed directly to the test.

#stress
str.arc.sync <- read.sync(file="ARC.STR.input", gen=c(0, 1, 0, 1, 0, 1), repl=c(1, 1, 2, 2, 3,3))
str.arc.cov <- coverage(str.arc.sync,gen=c(0, 1, 0, 1, 0, 1), repl=c(1, 1, 2, 2, 3,3))
str.arc.af <- af(str.arc.sync,gen=c(0, 1, 0, 1, 0, 1), repl=c(1, 1, 2, 2, 3,3))

#control
con.arc.sync <- read.sync(file="ARC.CON.input", gen=c(0, 1, 0, 1, 0, 1), repl=c(1, 1, 2, 2, 3,3))
con.arc.cov <- coverage(con.arc.sync,gen=c(0, 1, 0, 1, 0, 1), repl=c(1, 1, 2, 2, 3,3))
con.arc.af <- af(con.arc.sync,gen=c(0, 1, 0, 1, 0, 1), repl=c(1, 1, 2, 2, 3,3))
```

#### Running adapted CMH test
The adapted CMH is a replicated test of allele frequency change — it asks whether allele frequencies shifted consistently across all 3 IS→treatment replicate pairs. **The parameters matter a lot here:**
- Ne=rep(250, 3) — effective population size of 250 per replicate, used to correct for genetic drift
  - estimated Ne to be around 400 because this is 1000x less than starting pool size of 400,000 larv per bucket and 100x less than 40,000 larv per initial sample
- poolSize=rep(c(40000, 10000), ...) — pool sizes alternating between IS (40,000 larvae) and treatment (40,000 larvae), used for appropriate variance correction
- MeanStart=TRUE — uses the mean IS allele frequency across replicates as the baseline rather than each IS independently
- mincov=20 — excludes loci with coverage below 20x in any pool, preventing low-coverage noise from driving false positives
```{r}
#stress
str.arc.pvals <- adapted.cmh.test(freq=str.arc.af, coverage=str.arc.cov,
                                  Ne=rep(400, 3), gen=c(0, 1, 0, 1, 0, 1), repl=c(1, 1, 2, 2, 3,3),
                                  poolSize=rep(c(40000,40000), ncol(str.arc.af)/2), MeanStart = TRUE, mincov =20)

pstrdf <- as.data.table(cbind(row.names(str.arc.af),str.arc.pvals))
colnames(pstrdf) <- c("Locus","PVAL")
pstrdf$GROUP <- "PSTR"

#control
con.arc.pvals <- adapted.cmh.test(freq=con.arc.af, coverage=con.arc.cov,
                                  Ne=rep(400, 3), gen=c(0, 1, 0, 1, 0, 1), repl=c(1, 1, 2, 2, 3,3),
                                  poolSize=rep(c(40000,40000), ncol(con.arc.af)/2), MeanStart = TRUE, mincov =20)

pcondf <- as.data.table(cbind(row.names(con.arc.af),con.arc.pvals))
colnames(pcondf) <- c("Locus","PVAL")
pcondf$GROUP <- "PCON"

# build p value
ARC.pval.table <- bind_rows(pcondf, pstrdf)
ARC.pval.table <- ARC.pval.table %>%
  separate(Locus, into = c("CHROM", "BP"), sep = "\\.(?=[^.]+$)")
ARC.pval.table$PVAL <- as.numeric(ARC.pval.table$PVAL)
```



## NEH
First normalize with initial sample coverage. The core logic is that the internal standard (IS) pools (the inital samples) act as a reference to normalize and cap coverage, ensuring that outlier detection downstream isn't driven by coverage artifacts rather than true allele frequency signals.
```{bash, eval=FALSE}
source activate CADO

#sum up initial samples
bash ../Scripts/sum.sh <(cut -f1,2,3,10-12,19-21,28-30 NEH.dp25.cov.sync) > NEH.IS.sync
# makes extra columns with the sum data
paste NEH.IS.sync <(cut -f4 NEH.IS.sync) <(cut -f4 NEH.IS.sync)  > NEH.ISsum3.sync

# look at what col number to put in the compute average step next 
mawk 'NR==1 { for(i=1; i<=NF; i++) print i, $i }' NEH.dp25.cov.sync
mawk -f ../Scripts/add_cov_sync.txt <(cut -f10-12,19-21,28-30 --complement NEH.dp25.cov.sync) > NEH.sansIS.cov.sync
mawk 'NR==1 { for(i=1; i<=NF; i++) print i, $i }' NEH.sansIS.cov.sync # last column is 27 (mean coverage w/o initial samples)

# compute average coverage on sync file minus initial samples
AV_COV=$(mawk -f ../Scripts/add_cov_sync.txt <(cut -f10-12,19-21,28-30 --complement NEH.dp25.cov.sync) | mawk '{sum=sum+$27} END {print sum/NR}')
echo $AV_COV

mawk -f ../Scripts/add_cov_sync.txt NEH.ISsum3.sync | mawk -v x=$AV_COV '$7 >= x' | mawk '!/CHR/' | cut -f1-6 > NEH.ISsum3nh.sync

mawk -f ../Scripts/add_cov_sync_IS.txt NEH.ISsum3.sync | mawk -v x=$AV_COV '$7 >= x'> NEH.ISsum3nh.cov.sync

# apply coverage filter back to main sync

#first check col number to apply filter to
paste NEH.dp25.cov.sync <(mawk -f ../Scripts/add_cov_sync.txt NEH.ISsum3.sync ) > ISnormNEH.cov.sync
mawk 'NR==1 { for(i=1; i<=NF; i++) print i, $i }' ISnormNEH.cov.sync # col 42 (mean cov normalized by initial samples)

# apply initial sample coverage filter
paste NEH.dp25.cov.sync <(mawk -f ../Scripts/add_cov_sync.txt NEH.ISsum3.sync ) | mawk -v x=$AV_COV '$42 >= x' | cut -f1-33 > temp.sync

# NEH samples normalized by initial sample coverage stored in *COV.sync
mv temp.sync NEH.dp25.COV.sync
#cp NEH.dp25.cov.sync NEH.dp25.COV.sync

# cap coverage at IS maxiumum

#remove inital samples and remove old cov stats and recalculates the coverage with only treatment pools
mawk -f ../Scripts/add_cov_sync.txt <(cut -f10-12,19-21,28- --complement NEH.dp25.COV.sync) > temp.cov

# pastes pop coverage with minimum inital sample coverage (col 7) to last col (25) of temp.cov and compare treatemtn pool mean cov ($24) to IS min cov ($25)
paste temp.cov <(cut -f7 NEH.ISsum3nh.cov.sync) | mawk -v OFS='\t' '{if ($24 > $25) {$24=$25; print $0} else {print $0}}' | cut -f1-24> NEH.dp25.COV
```
The final output is NEH.dp25.COV, a fully processed, coverage-normalized sync ready for outlier detection.


```{bash, eval=FALSE}
source activate random_draw

# subsampling the internal standard 
#python ../scripts/sub_sample.py NEH.dp25.COV NEH.ISsum3nh.sync NEH.ISs.sync
python ../Scripts/sub_sample.py NEH.ISsum3nh.cov.sync NEH.ISsum3nh.sync NEH.ISs.sync

# add heading
cat <(echo -e "CHROM\tPOS\tREF\tIS_RS1\tIS_RS2\tIS_RS3") NEH.ISs.sync > NEH.ISsum.sync

# building treatment specific input files (interleaved with IS columns)
#these IS columns will serve as the null expectation during outlier detection

# Control columns - will need to update columns based on data
paste <(cut -f1,2,3,4 NEH.ISsum.sync) <(cut -f4 NEH.dp25.COV.sync) \
      <(cut -f5 NEH.ISsum.sync) <(cut -f5 NEH.dp25.COV.sync) \
      <(cut -f6 NEH.ISsum.sync) <(cut -f6 NEH.dp25.COV.sync) \
      <(cut -f4 NEH.ISsum.sync) <(cut -f13 NEH.dp25.COV.sync) \
      <(cut -f5 NEH.ISsum.sync) <(cut -f14 NEH.dp25.COV.sync) \
      <(cut -f6 NEH.ISsum.sync) <(cut -f15 NEH.dp25.COV.sync) \
      <(cut -f4 NEH.ISsum.sync) <(cut -f22 NEH.dp25.COV.sync) \
      <(cut -f5 NEH.ISsum.sync) <(cut -f23 NEH.dp25.COV.sync) \
      <(cut -f6 NEH.ISsum.sync) <(cut -f24 NEH.dp25.COV.sync)> NEH.CON.sync

head -1 NEH.CON.sync
tail -n +2 NEH.CON.sync > NEH.CON.input

# Stress columns - will need to update columns based on data
paste <(cut -f1,2,3,4 NEH.ISsum.sync)  <(cut -f7 NEH.dp25.COV.sync) \
      <(cut -f5 NEH.ISsum.sync) <(cut -f8 NEH.dp25.COV.sync) \
      <(cut -f6 NEH.ISsum.sync) <(cut -f9 NEH.dp25.COV.sync) \
      <(cut -f4 NEH.ISsum.sync) <(cut -f16 NEH.dp25.COV.sync) \
      <(cut -f5 NEH.ISsum.sync) <(cut -f17 NEH.dp25.COV.sync) \
      <(cut -f6 NEH.ISsum.sync) <(cut -f18 NEH.dp25.COV.sync) \
      <(cut -f4 NEH.ISsum.sync) <(cut -f25 NEH.dp25.COV.sync) \
      <(cut -f5 NEH.ISsum.sync) <(cut -f26 NEH.dp25.COV.sync) \
      <(cut -f6 NEH.ISsum.sync) <(cut -f27 NEH.dp25.COV.sync)> NEH.STR.sync

head -1 NEH.STR.sync
tail -n +2 NEH.STR.sync > NEH.STR.input
```

### Calculate P-values - NEH
#### Modified CMH - NEH population

Reads in the interleaved IS+treatment sync files into R and runs an adapted Cochran-Mantel-Haenszel (CMH) test to identify loci where allele frequencies shift significantly between initial samples and treatment pools.

Reading and parsing files - read.sync() requires a headerless input file
```{r}
# gen assigns each column a "generation," where 0 = IS (baseline) and 1 = treatment pool
# repl  pairs each IS replicate with its corresponding treatment pool as replicate 1, 2, and 3
# coverage() and af() extract per-locus read depth and allele frequencies from the parsed sync object, which are passed directly to the test.

#stress
str.neh.sync <- read.sync(file="NEH.STR.input", gen=c(0, 1, 0, 1, 0, 1, 0, 1, 0, 1, 0, 1, 0, 1, 0, 1, 0, 1), repl=c(1, 1, 2, 2, 3,3, 4,4, 5,5, 6,6, 7, 7, 8,8, 9,9))
str.neh.cov <- coverage(str.neh.sync, gen=c(0, 1, 0, 1, 0, 1, 0, 1, 0, 1, 0, 1, 0, 1, 0, 1, 0, 1), repl=c(1, 1, 2, 2, 3,3, 4,4, 5,5, 6,6, 7, 7, 8,8, 9,9))
str.neh.af <- af(str.neh.sync, gen=c(0, 1, 0, 1, 0, 1, 0, 1, 0, 1, 0, 1, 0, 1, 0, 1, 0, 1), repl=c(1, 1, 2, 2, 3,3, 4,4, 5,5, 6,6, 7, 7, 8,8, 9,9))

#control
con.neh.sync <- read.sync(file="NEH.CON.input", gen=c(0, 1, 0, 1, 0, 1, 0, 1, 0, 1, 0, 1, 0, 1, 0, 1, 0, 1), repl=c(1, 1, 2, 2, 3,3, 4,4, 5,5, 6,6, 7, 7, 8,8, 9,9))
con.neh.cov <- coverage(con.neh.sync, gen=c(0, 1, 0, 1, 0, 1, 0, 1, 0, 1, 0, 1, 0, 1, 0, 1, 0, 1), repl=c(1, 1, 2, 2, 3,3, 4,4, 5,5, 6,6, 7, 7, 8,8, 9,9))
con.neh.af <- af(con.neh.sync, gen=c(0, 1, 0, 1, 0, 1, 0, 1, 0, 1, 0, 1, 0, 1, 0, 1, 0, 1), repl=c(1, 1, 2, 2, 3,3, 4,4, 5,5, 6,6, 7, 7, 8,8, 9,9))
```

#### Running adapted CMH test
The adapted CMH is a replicated test of allele frequency change — it asks whether allele frequencies shifted consistently across all 3 IS→treatment replicate pairs. **The parameters matter a lot here:**
- Ne=rep(250, 3) — effective population size of 250 per replicate, used to correct for genetic drift
  - estimated Ne to be around 400 because this is 1000x less than starting pool size of 400,000 larv per bucket and 100x less than 40,000 larv per initial sample
- poolSize=rep(c(40000, 10000), ...) — pool sizes alternating between IS (40,000 larvae) and treatment (40,000 larvae), used for appropriate variance correction
- MeanStart=TRUE — uses the mean IS allele frequency across replicates as the baseline rather than each IS independently
- mincov=20 — excludes loci with coverage below 20x in any pool, preventing low-coverage noise from driving false positives
```{r}
#stress
str.neh.pvals <- adapted.cmh.test(freq=str.neh.af, coverage=str.neh.cov,
                                  Ne=rep(400, 9), gen=c(0, 1, 0, 1, 0, 1, 0, 1, 0, 1, 0, 1, 0, 1, 0, 1, 0, 1), 
                                  repl=c(1, 1, 2, 2, 3,3, 4,4, 5,5, 6,6, 7, 7, 8,8, 9,9),
                                  poolSize=rep(c(40000,20000), ncol(str.neh.af)/2), MeanStart = TRUE, mincov =20)

pstrdf <- as.data.table(cbind(row.names(str.neh.af),str.neh.pvals))
colnames(pstrdf) <- c("Locus","PVAL")
pstrdf$GROUP <- "PSTR"

#control
con.neh.pvals <- adapted.cmh.test(freq=con.neh.af, coverage=con.neh.cov,
                                  Ne=rep(400, 9), gen=c(0, 1, 0, 1, 0, 1, 0, 1, 0, 1, 0, 1, 0, 1, 0, 1, 0, 1), 
                                  repl=c(1, 1, 2, 2, 3,3, 4,4, 5,5, 6,6, 7, 7, 8,8, 9,9),
                                  poolSize=rep(c(40000,20000), ncol(con.neh.af)/2), MeanStart = TRUE, mincov =20)

pcondf <- as.data.table(cbind(row.names(con.neh.af),con.neh.pvals))
colnames(pcondf) <- c("Locus","PVAL")
pcondf$GROUP <- "PCON"

# build p value
NEH.pval.table <- bind_rows(pcondf, pstrdf)
NEH.pval.table <- NEH.pval.table %>%
  separate(Locus, into = c("CHROM", "BP"), sep = "\\.(?=[^.]+$)")
NEH.pval.table$PVAL <- as.numeric(NEH.pval.table$PVAL)
```

## MV 
First normalize with initial sample coverage. The core logic is that the internal standard (IS) pools (the inital samples) act as a reference to normalize and cap coverage, ensuring that outlier detection downstream isn't driven by coverage artifacts rather than true allele frequency signals.
```{bash, eval=FALSE}
source activate CADO

#sum up initial samples
bash ../Scripts/sum.sh <(cut -f1,2,3,10-12,19-21,28-30 MV.dp25.cov.sync) > MV.IS.sync
# makes extra columns with the sum data
paste MV.IS.sync <(cut -f4 MV.IS.sync) <(cut -f4 MV.IS.sync)  > MV.ISsum3.sync

# look at what col number to put in the compute average step next 
mawk 'NR==1 { for(i=1; i<=NF; i++) print i, $i }' MV.dp25.cov.sync
mawk -f ../Scripts/add_cov_sync.txt <(cut -f10-12,19-21,28-30 --complement MV.dp25.cov.sync) > MV.sansIS.cov.sync
mawk 'NR==1 { for(i=1; i<=NF; i++) print i, $i }' MV.sansIS.cov.sync # last column is 27 (mean coverage w/o initial samples)

# compute average coverage on sync file minus initial samples
AV_COV=$(mawk -f ../Scripts/add_cov_sync.txt <(cut -f10-12,19-21,28-30 --complement MV.dp25.cov.sync) | mawk '{sum=sum+$27} END {print sum/NR}')
echo $AV_COV

mawk -f ../Scripts/add_cov_sync.txt MV.ISsum3.sync | mawk -v x=$AV_COV '$7 >= x' | mawk '!/CHR/' | cut -f1-6 > MV.ISsum3nh.sync

mawk -f ../Scripts/add_cov_sync_IS.txt MV.ISsum3.sync | mawk -v x=$AV_COV '$7 >= x'> MV.ISsum3nh.cov.sync

# apply coverage filter back to main sync

#first check col number to apply filter to
paste MV.dp25.cov.sync <(mawk -f ../Scripts/add_cov_sync.txt MV.ISsum3.sync ) > ISnormMV.cov.sync
mawk 'NR==1 { for(i=1; i<=NF; i++) print i, $i }' ISnormMV.cov.sync # col 42 (mean cov normalized by initial samples)

# apply initial sample coverage filter
paste MV.dp25.cov.sync <(mawk -f ../Scripts/add_cov_sync.txt MV.ISsum3.sync ) | mawk -v x=$AV_COV '$42 >= x' | cut -f1-33 > temp.sync

# MV samples normalized by initial sample coverage stored in *COV.sync
mv temp.sync MV.dp25.COV.sync
#cp MV.dp25.cov.sync MV.dp25.COV.sync

# cap coverage at IS maxiumum

#remove inital samples and remove old cov stats and recalculates the coverage with only treatment pools
mawk -f ../Scripts/add_cov_sync.txt <(cut -f10-12,19-21,28- --complement MV.dp25.COV.sync) > temp.cov

# pastes pop coverage with minimum inital sample coverage (col 7) to last col (25) of temp.cov and compare treatemtn pool mean cov ($24) to IS min cov ($25)
paste temp.cov <(cut -f7 MV.ISsum3nh.cov.sync) | mawk -v OFS='\t' '{if ($24 > $25) {$24=$25; print $0} else {print $0}}' | cut -f1-24> MV.dp25.COV
```
The final output is MV.dp25.COV, a fully processed, coverage-normalized sync ready for outlier detection.


```{bash, eval=FALSE}
source activate random_draw

# subsampling the internal standard 
#python ../scripts/sub_sample.py MV.dp25.COV MV.ISsum3nh.sync MV.ISs.sync
python ../Scripts/sub_sample.py MV.ISsum3nh.cov.sync MV.ISsum3nh.sync MV.ISs.sync

# add heading
cat <(echo -e "CHROM\tPOS\tREF\tIS_RS1\tIS_RS2\tIS_RS3") MV.ISs.sync > MV.ISsum.sync

# building treatment specific input files (interleaved with IS columns)
#these IS columns will serve as the null expectation during outlier detection

# Control columns - will need to update columns based on data
paste <(cut -f1,2,3,4 MV.ISsum.sync) <(cut -f4 MV.dp25.COV.sync) \
      <(cut -f5 MV.ISsum.sync) <(cut -f5 MV.dp25.COV.sync) \
      <(cut -f6 MV.ISsum.sync) <(cut -f6 MV.dp25.COV.sync) \
      <(cut -f4 MV.ISsum.sync) <(cut -f13 MV.dp25.COV.sync) \
      <(cut -f5 MV.ISsum.sync) <(cut -f14 MV.dp25.COV.sync) \
      <(cut -f6 MV.ISsum.sync) <(cut -f15 MV.dp25.COV.sync) \
      <(cut -f4 MV.ISsum.sync) <(cut -f22 MV.dp25.COV.sync) \
      <(cut -f5 MV.ISsum.sync) <(cut -f23 MV.dp25.COV.sync) \
      <(cut -f6 MV.ISsum.sync) <(cut -f24 MV.dp25.COV.sync)> MV.CON.sync

head -1 MV.CON.sync
tail -n +2 MV.CON.sync > MV.CON.input

# Stress columns - will need to update columns based on data
paste <(cut -f1,2,3,4 MV.ISsum.sync)  <(cut -f7 MV.dp25.COV.sync) \
      <(cut -f5 MV.ISsum.sync) <(cut -f8 MV.dp25.COV.sync) \
      <(cut -f6 MV.ISsum.sync) <(cut -f9 MV.dp25.COV.sync) \
      <(cut -f4 MV.ISsum.sync) <(cut -f16 MV.dp25.COV.sync) \
      <(cut -f5 MV.ISsum.sync) <(cut -f17 MV.dp25.COV.sync) \
      <(cut -f6 MV.ISsum.sync) <(cut -f18 MV.dp25.COV.sync) \
      <(cut -f4 MV.ISsum.sync) <(cut -f25 MV.dp25.COV.sync) \
      <(cut -f5 MV.ISsum.sync) <(cut -f26 MV.dp25.COV.sync) \
      <(cut -f6 MV.ISsum.sync) <(cut -f27 MV.dp25.COV.sync)> MV.STR.sync

head -1 MV.STR.sync
tail -n +2 MV.STR.sync > MV.STR.input
```

### Calculate P-values - MV
#### Modified CMH - MV population

Reads in the interleaved IS+treatment sync files into R and runs an adapted Cochran-Mantel-Haenszel (CMH) test to identify loci where allele frequencies shift significantly between initial samples and treatment pools.

Reading and parsing files - read.sync() requires a headerless input file
```{r}
# gen assigns each column a "generation," where 0 = IS (baseline) and 1 = treatment pool
# repl  pairs each IS replicate with its corresponding treatment pool as replicate 1, 2, and 3
# coverage() and af() extract per-locus read depth and allele frequencies from the parsed sync object, which are passed directly to the test.

#stress
str.mv.sync <- read.sync(file="MV.STR.input", gen=c(0, 1, 0, 1, 0, 1, 0, 1, 0, 1, 0, 1, 0, 1, 0, 1, 0, 1), repl=c(1, 1, 2, 2, 3,3, 4,4, 5,5, 6,6, 7, 7, 8,8, 9,9))
str.mv.cov <- coverage(str.mv.sync, gen=c(0, 1, 0, 1, 0, 1, 0, 1, 0, 1, 0, 1, 0, 1, 0, 1, 0, 1), repl=c(1, 1, 2, 2, 3,3, 4,4, 5,5, 6,6, 7, 7, 8,8, 9,9))
str.mv.af <- af(str.mv.sync, gen=c(0, 1, 0, 1, 0, 1, 0, 1, 0, 1, 0, 1, 0, 1, 0, 1, 0, 1), repl=c(1, 1, 2, 2, 3,3, 4,4, 5,5, 6,6, 7, 7, 8,8, 9,9))

#control
con.mv.sync <- read.sync(file="MV.CON.input", gen=c(0, 1, 0, 1, 0, 1, 0, 1, 0, 1, 0, 1, 0, 1, 0, 1, 0, 1), repl=c(1, 1, 2, 2, 3,3, 4,4, 5,5, 6,6, 7, 7, 8,8, 9,9))
con.mv.cov <- coverage(con.mv.sync, gen=c(0, 1, 0, 1, 0, 1, 0, 1, 0, 1, 0, 1, 0, 1, 0, 1, 0, 1), repl=c(1, 1, 2, 2, 3,3, 4,4, 5,5, 6,6, 7, 7, 8,8, 9,9))
con.mv.af <- af(con.mv.sync, gen=c(0, 1, 0, 1, 0, 1, 0, 1, 0, 1, 0, 1, 0, 1, 0, 1, 0, 1), repl=c(1, 1, 2, 2, 3,3, 4,4, 5,5, 6,6, 7, 7, 8,8, 9,9))
```

#### Running adapted CMH test
The adapted CMH is a replicated test of allele frequency change — it asks whether allele frequencies shifted consistently across all 3 IS→treatment replicate pairs. **The parameters matter a lot here:**
- Ne=rep(250, 3) — effective population size of 250 per replicate, used to correct for genetic drift
  - estimated Ne to be around 400 because this is 1000x less than starting pool size of 400,000 larv per bucket and 100x less than 40,000 larv per initial sample
- poolSize=rep(c(40000, 10000), ...) — pool sizes alternating between IS (40,000 larvae) and treatment (40,000 larvae), used for appropriate variance correction
- MeanStart=TRUE — uses the mean IS allele frequency across replicates as the baseline rather than each IS independently
- mincov=20 — excludes loci with coverage below 20x in any pool, preventing low-coverage noise from driving false positives
```{r}
#stress
str.mv.pvals <- adapted.cmh.test(freq=str.mv.af, coverage=str.mv.cov,
                                  Ne=rep(400, 9), gen=c(0, 1, 0, 1, 0, 1, 0, 1, 0, 1, 0, 1, 0, 1, 0, 1, 0, 1), 
                                  repl=c(1, 1, 2, 2, 3,3, 4,4, 5,5, 6,6, 7, 7, 8,8, 9,9),
                                  poolSize=rep(c(40000,40000), ncol(str.mv.af)/2), MeanStart = TRUE, mincov =20)

pstrdf <- as.data.table(cbind(row.names(str.mv.af),str.mv.pvals))
colnames(pstrdf) <- c("Locus","PVAL")
pstrdf$GROUP <- "PSTR"

#control
con.mv.pvals <- adapted.cmh.test(freq=con.mv.af, coverage=con.mv.cov,
                                  Ne=rep(400, 9), gen=c(0, 1, 0, 1, 0, 1, 0, 1, 0, 1, 0, 1, 0, 1, 0, 1, 0, 1), 
                                  repl=c(1, 1, 2, 2, 3,3, 4,4, 5,5, 6,6, 7, 7, 8,8, 9,9),
                                  poolSize=rep(c(40000,40000), ncol(con.mv.af)/2), MeanStart = TRUE, mincov =20)

pcondf <- as.data.table(cbind(row.names(con.mv.af),con.mv.pvals))
colnames(pcondf) <- c("Locus","PVAL")
pcondf$GROUP <- "PCON"

# build p value
MV.pval.table <- bind_rows(pcondf, pstrdf)
MV.pval.table <- MV.pval.table %>%
  separate(Locus, into = c("CHROM", "BP"), sep = "\\.(?=[^.]+$)")
MV.pval.table$PVAL <- as.numeric(MV.pval.table$PVAL)
```

#### convert p values to q values
This applies FDR (false discovery rate) correction to your CMH p-values for each population, converting raw p-values into q-values
```{r}
Qvalue_convert(ARC.pval.table) ->ARC.pv
Qvalue_convert(NEH.pval.table) ->NEH.pv
Qvalue_convert(MV.pval.table) ->MV.pv
```



## Outlier filtering - 10% FDR, 5% FDR and 1% FDR datasets - we'll use the 1% dataset
identify high-confidence candidate loci under selection due to stress across three populations
```{r}
pp1 <- rbind(ARC.pv,NEH.pv,MV.pv)

# Filter significant loci at FDR 10% threshold; also apply top 10% quantile filter within significant loci
alpha = 0.1
alpha2 =0.1


ppsig.ARC.1 <- Significant_subset(ARC.pv,alpha,alpha2)
ppsig.NEH.1 <- Significant_subset(NEH.pv,alpha,alpha2)
ppsig.MV.1 <- Significant_subset(MV.pv,alpha,alpha2)

ppsig.ARC.1$POP <- "ARC"
ppsig.NEH.1$POP <- "NEH"
ppsig.MV.1$POP <- "MV"

ppsig3.FDR10 <- bind_rows(ppsig.ARC.1,ppsig.NEH.1,ppsig.MV.1)

#all_sig.STR<- subset(ppsig3.FDR10, Sig.STR == "TRUE") %>% group_by(SNP) %>% filter(n()>2)
#all_sig.STR$Sig.CON<- FALSE

#write.table(all_sig.STR, "Sig.Loci.3", sep="\t", row.names = FALSE, quote = FALSE)
```


```{r}
# Filter significant loci at FDR 5% threshold; also apply top 10% quantile filter within significant loci
alpha = 0.05
#NEH.alpha = alpha * 2
alpha2 =0.1

ppsig.ARC.05 <- Significant_subset(ARC.pv,alpha,alpha2)
ppsig.NEH.05 <- Significant_subset(NEH.pv,alpha,alpha2)
ppsig.MV.05 <- Significant_subset(MV.pv,alpha,alpha2)

ppsig.ARC.05$POP <- "ARC"
ppsig.NEH.05$POP <- "NEH"
ppsig.MV.05$POP <- "MV"

ppsig3.FDR05 <- rbind(ppsig.ARC.05,ppsig.NEH.05,ppsig.MV.05)

#multi_sig.STR<- subset(ppsig3.FDR05, Sig.STR == "TRUE") %>% group_by(SNP) %>% filter(n()>1)
#multi_sig.STR$Sig.CON <- FALSE

#multi_sig.CA<- subset(ppsig3.FDR05, QCA < NEH.alpha) %>% group_by(SNP) %>% filter(n()>1)
#multi_sig.CASE<- subset(ppsig3.FDR05, QCASE < NEH.alpha) %>% group_by(SNP) %>% filter(n()>1)
#multi_sig.SE<- subset(ppsig3.FDR05, QSE < NEH.alpha) %>% group_by(SNP) %>% filter(n()>1)

#multi_sig <- rbind(ppsig3.FDR05 %>% group_by(SNP) %>% filter(n()>1))
#multi_sig <- rbind(multi_sig.CA,multi_sig.CASE,multi_sig.SE)

#write.table(multi_sig.STR, "Sig.Loci.2", sep="\t", row.names = FALSE, quote = FALSE)
```

```{r}

#Filter significant loci at 1% FDR threshold; and take the top 10% 
alpha = 0.01
alpha2 =0.1
#NEH.alpha = alpha *2

ppsig.ARC.01 <- Significant_subset(ARC.pv,alpha,alpha2)
ppsig.NEH.01 <- Significant_subset(NEH.pv,alpha,alpha2)
ppsig.MV.01 <- Significant_subset(MV.pv,alpha,alpha2)
```

```{r}
conpp1 <- subset(pp1, QCON < alpha)

ppsig.ARC.01$POP <- "ARC"
ppsig.NEH.01$POP <- "NEH"
ppsig.MV.01$POP <- "MV"

ppsig3.FDR01 <- rbind(ppsig.ARC.01,ppsig.NEH.01,ppsig.MV.01)

write.table(ppsig3.FDR01, "Sig.Loci.FDR01.1", sep="\t", row.names = FALSE, quote = FALSE)
write.table(ppsig.ARC.01, "Sig.Loci.FDR01.ARC", sep="\t", row.names = FALSE, quote = FALSE)
write.table(ppsig.NEH.01, "Sig.Loci.FDR01.NEH", sep="\t", row.names = FALSE, quote = FALSE)
write.table(ppsig.MV.01, "Sig.Loci.FDR01.MV", sep="\t", row.names = FALSE, quote = FALSE)
```

> nrow(ppsig.ARC.01)
[1] 276
> nrow(ppsig.NEH.01)
[1] 2216
> nrow(ppsig.MV.01)
[1] 4180

** I have filtered down to FDR of 1% **

#### extra filtering steps that Jon
I don't think I need these - they were used when Jon was building his all blocks total loci file. 
For individual population analysis I don't think I need to include this addition filter. 
```{r}
# don't need this extra filtering - this would be used if I was combining the populations and analyzing them together
# multi_sig.STR.01<- subset(ppsig3.FDR01, Sig.STR == "TRUE") %>% group_by(SNP) %>% filter(n()>1)
# multi_sig.STR.01$Sig.CON <- FALSE
# 
# write.table(multi_sig.STR.01, "Sig.Loci.FDR01.2", sep="\t", row.names = FALSE, quote = FALSE)
# 
# alpha = 0.01
# alpha2 =0.1
# NEH.alpha = alpha * 2
# 
# 
# ppsig.ARC <- Significant_subset(ARC.pv,alpha,alpha2)
#ppsig.ARC <- subset(ppsig.ARC, Sig.STR == TRUE & QSTR < quantile(ppsig.ARC$QSTR, na.rm = TRUE, probs = 0.01)
#                     )
# ppsig.NEH <- Significant_subset(NEH.pv,NEH.alpha,alpha2)
# ppsig.NEH <- subset(ppsig.NEH, Sig.STR == TRUE & QSTR < quantile(ppsig.NEH$QSTR, na.rm = TRUE, probs = 0.01)
#                     )
# ppsig.MV <- Significant_subset(MV.pv,alpha,alpha2)
#ppsig.MV <- subset(ppsig.MV, Sig.STR == TRUE & QSTR < quantile(ppsig.MV$QSTR, na.rm = TRUE, probs = 0.01)
#                     )
# 
# ppsig.ARC$POP <- "ARC"
# ppsig.NEH$POP <- "NEH"
# ppsig.MV$POP <- "MV"
# 
# singleton <- read.table("singleton.loci",header=TRUE)
# 
# ppsig1.s <- semi_join(rbind(ppsig.ARC,ppsig.NEH,ppsig.MV),singleton, by = "SNP")
```


## Control filtering by population
```{r}
# ARC
conpp.ARC <- subset(ARC.pv, QCON < 0.001)
allcon.ARC <- subset(conpp.ARC, QCON / QSTR < 10)
write.table(allcon.ARC, "ARC.ALL.CON.SNPS", sep="\t", row.names=FALSE, quote=FALSE)

# NEH
conpp.NEH <- subset(NEH.pv, QCON < 0.001)
allcon.NEH <- subset(conpp.NEH, QCON / QSTR < 10)
write.table(allcon.NEH, "NEH.ALL.CON.SNPS", sep="\t", row.names=FALSE, quote=FALSE)

# MV
conpp.MV <- subset(MV.pv, QCON < 0.001)
allcon.MV <- subset(conpp.MV, QCON / QSTR < 10)
write.table(allcon.MV, "MV.ALL.CON.SNPS", sep="\t", row.names=FALSE, quote=FALSE)
```

```{r}
total.sig.ARC <- anti_join(ppsig.ARC.01, allcon.ARC, by = "SNP")
total.sig.NEH <- anti_join(ppsig.NEH.01, allcon.NEH, by = "SNP")
total.sig.MV  <- anti_join(ppsig.MV.01,  allcon.MV,  by = "SNP")

write.table(total.sig.ARC, "Total.Significant.uf.Loci.ARC", sep="\t", row.names=FALSE, quote=FALSE)
write.table(total.sig.NEH, "Total.Significant.uf.Loci.NEH", sep="\t", row.names=FALSE, quote=FALSE)
write.table(total.sig.MV,  "Total.Significant.uf.Loci.MV",  sep="\t", row.names=FALSE, quote=FALSE)
```   

```{bash}
source activate CADO

for POP in ARC NEH MV; do

  # Build per-population significant loci BED
  cut -f2,3 Total.Significant.uf.Loci.${POP} | tail -n +2 | sort -k1,2 | uniq \
    | mawk '{print $1 "\t" $2-1 "\t" $2}' \
    | bedtools sort -i - > CASE.Total.Significant.uf.loci.${POP}.bed

  # Build per-population CON loci gene list
  bedtools intersect -wb \
    -a <(mawk '{print $2"\t"$3-1"\t"$3}' ${POP}.ALL.CON.SNPS) \
    -b /home/Genomic_Resources/C_virginica/bed_and_GFF/sorted.ref3.0.gene.bed \
    | grep -oh "gene=LOC.*;g" | sed 's/gene=//g' | sed 's/;g//g' \
    | sort | uniq > CON.LOCI.${POP}

  # Build per-population CON filter BED
  bedtools intersect -wb \
    -a <(mawk '{print $2"\t"$3-1"\t"$3}' ${POP}.ALL.CON.SNPS) \
    -b /home/Genomic_Resources/C_virginica/bed_and_GFF/sorted.ref3.0.gene.bed \
    | cut -f4- | bedtools sort | bedtools merge > con.loci.filter.${POP}.bed

  # Build per-population ToFilter file
  cat <(echo -e "SNP\tCHROM\tBP") \
    <(bedtools intersect \
        -a CASE.Total.Significant.uf.loci.${POP}.bed \
        -b con.loci.filter.${POP}.bed \
      | mawk '{print $1"_"$3"\t"$1"\t"$3}') \
    > ToFilter.Con.Loci.${POP}.txt

done
```


```{r}
library(dplyr)

pops <- c("ARC", "NEH", "MV")

for (pop in pops) {

  con_filter <- read.table(paste0("ToFilter.Con.Loci.", pop, ".txt"), header = TRUE)

  total.sig <- get(paste0("total.sig.", pop))

  # Remove contaminating loci
  total.sig <- unique(anti_join(total.sig, con_filter, by = "SNP"))

  write.table(total.sig,
              paste0("Total.Significant.Loci.", pop),
              sep = "\t", row.names = FALSE,  quote = FALSE)
}
```


#### Can do BED conversion here and GO enrichment, but skipping for now



# Manhattan Plots

## Per population Outlier Manhattan Plot Composite
```{r}
library(dplyr)
library(ggplot2)
library(patchwork)

make_manhattan_plot <- function(pv_data, sig_data, con_filter, allcon, pop_name,
                                outlier_color, y_limit = NULL) {
  
  pv_data <- pv_data %>%
    mutate(
      Group = case_when(
        SNP %in% con_filter$SNP ~ "Control Treatment Filtered",
        SNP %in% allcon$SNP ~ "Control Treatment Filtered",
        TRUE ~ "Neutral"
      )
    )
  
  man <- ggman(
    pv_data,
    pvalue = "QSTR",
    chr = "CHR",
    snp = "SNP",
    bp = "BP",
    relative.positions = TRUE,
    sigLine = NA,
    pointSize = 1.5,
    title = "",
    alpha = 0.5
  )
  
  dfm <- man[[1]]
  dfm$chrom_alt <- factor(dfm$chrom_alt, levels = c(0, 1))
  
  dfmo <- semi_join(dfm, sig_data, by = "SNP")
  dfmo$Group <- "Outlier"
  
  dfmsplit <- split(dfm, dfm$chrom)
  xbreaks <- sapply(dfmsplit, function(x) {
    midpoint <- length(x$index) / 2
    if (midpoint < 1) midpoint <- 1
    x$index[midpoint]
  })
  
  y_max <- if (is.null(y_limit)) max(dfm$marker + 1, na.rm = TRUE) else y_limit
  
  p <- ggplot(dfm, aes(x = index, y = marker, colour = as.factor(chrom_alt))) +
    geom_point(aes(alpha = marker, shape = Group, size = marker)) +
    geom_point(
      data = dfmo,
      aes(x = index, y = marker, shape = Group, size = marker),
      color = outlier_color,
      fill = outlier_color
    ) +
    scale_x_continuous(
      breaks = xbreaks,
      labels = names(xbreaks),
      expand = c(0, 0),
      limits = c(0, max(dfm$index) + 10)
    ) +
    guides(colour = FALSE, alpha = FALSE, fill = FALSE, size = FALSE) +
    scale_y_continuous(expand = c(0, 0), limits = c(0, y_max)) +
    scale_size_continuous(range = c(0.25, 3)) +
    scale_alpha_continuous(range = c(0.05, 0.8), limits = c(0, 50)) +
    ggtitle(pop_name) +
    labs(x = "Chromosome", y = expression(-Log10(q-value))) +
    theme_classic() +
    scale_shape_manual(values = c(15, 21, 16, 17)) +
    scale_color_manual(values = c("grey", "dark grey"), guide = FALSE) +
    guides(shape = guide_legend(override.aes = list(color = "grey", size = 5, linetype = 0)))
  
  list(plot = p, dfm = dfm, dfmo = dfmo)
}

```

```{r}
arc_con_filter <- read.table("ToFilter.Con.Loci.ARC.txt", header = TRUE)
neh_con_filter <- read.table("ToFilter.Con.Loci.NEH.txt", header = TRUE)
mv_con_filter  <- read.table("ToFilter.Con.Loci.MV.txt",  header = TRUE)

#load in control filtered 1% FDR loci for each population
total.sig.ARC.filt <- read.table("Total.Significant.Loci.ARC", header = TRUE)
total.sig.NEH.filt <- read.table("Total.Significant.Loci.NEH", header = TRUE)
total.sig.MV.filt <- read.table("Total.Significant.Loci.MV", header = TRUE)
```

> nrow(total.sig.ARC)
[1] 276
> nrow(total.sig.ARC.filt)
[1] 268 control filtered outliers (p<0.01)
> nrow(total.sig.NEH)
[1] 2216
> nrow(total.sig.NEH.filt)
[1] 1321 control filtered outliers (p<0.01)
> nrow(total.sig.MV)
[1] 4180
> nrow(total.sig.MV.filt)
[1] 2627 control filtered outliers (p<0.01)


```{r}
arc.h <- subset(total.sig.ARC.filt, Sig.STR == TRUE)
neh.h <- subset(total.sig.NEH.filt, Sig.STR == TRUE)
mv.h  <- subset(total.sig.MV.filt,  Sig.STR == TRUE)

arc_res <- make_manhattan_plot(
  pv_data = ARC.pv,
  sig_data = arc.h,
  con_filter = arc_con_filter,
  allcon = allcon.ARC,
  pop_name = "ARC",
  outlier_color = "#3F8F7A"
)

neh_res <- make_manhattan_plot(
  pv_data = NEH.pv,
  sig_data = neh.h,
  con_filter = neh_con_filter,
  allcon = allcon.NEH,
  pop_name = "NEH",
  outlier_color = "#D8848A",
  y_limit = 60
)

mv_res <- make_manhattan_plot(
  pv_data = MV.pv,
  sig_data = mv.h,
  con_filter = mv_con_filter,
  allcon = allcon.MV,
  pop_name = "MV",
  outlier_color = "#5F7BAA"
)

```


```{r}
png(filename = "ManhattanPlots/All_Populations_Manhattan_newfilt_2.png", 
    units = "in", width = 10, height = 6,  res = 300, bg = "transparent")
arc_res$plot / neh_res$plot / mv_res$plot +
  plot_annotation(tag_levels = "A") +
  plot_layout(guides = "collect") 
dev.off()
```



## ARC - Outlier Manhattan Plot
```{r}
arc_con_filter <- read.table("ToFilter.Con.Loci.ARC.txt", header = TRUE)
```

```{r, warning =FALSE, eval = FALSE}
arc.h <- subset(total.sig.ARC, Sig.STR == TRUE)
arc.c <- subset(ARC.pv, QCON < 0.1)


ARC.pp <- ARC.pv %>%
  mutate(
    Group = case_when(
      #SNP %in% singleton$SNP ~ "Singleton",
      SNP %in% arc.c$SNP ~ "Control Treatment Filtered",
      SNP %in% allcon.ARC$SNP ~ "Control Treatment Filtered",
      SNP %in% arc_con_filter$SNP ~ "Control Treatment Filtered",
      TRUE ~ "Shared SNP"
    )
  )

man <-ggman(ARC.pp, pvalue="QSTR", chr="CHR", snp="SNP",bp="BP", relative.positions = TRUE , sigLine = NA,
            pointSize=1.5,title = "", alpha = 0.5)

dfm <- man[[1]]

dfm$chrom_alt <- factor(dfm$chrom_alt, levels=c(0,1))

dfmo <- semi_join(dfm, arc.h, by = "SNP")
dfmo$Group <- "Outlier"

dfmsplit <- split(dfm, dfm$chrom)
xbreaks <- sapply(dfmsplit,function(x) {
    midpoint <- length(x$index)/2
    if(midpoint <1) midpoint <- 1
    return(x$index[midpoint])
})
arc.STR <-ggplot(dfm, aes(x= index, y=marker, colour = as.factor(chrom_alt)))+
  geom_point(aes(alpha=marker, shape=Group, size = marker))+  
  geom_point(data=dfmo, fill=cbPaletteSmall3[1], aes(x=index, y=marker, shape=Group, size=marker),  color = cbPaletteSmall3[1])+
  scale_x_continuous(breaks = xbreaks, labels = names(xbreaks),expand = c(0,0),limits = c(0,max(dfm$index)+10)) +
  guides(colour = FALSE,alpha=FALSE, fill = FALSE, size = FALSE) +
  scale_y_continuous(expand = c(0,0),limits=c(0,max(dfm$marker+1,na.rm=TRUE)))+
  scale_size_continuous(range = c(0.25,3)) +
  scale_alpha_continuous(range=c(0.05,0.8), limits= c(0,50))+
  ggtitle("ARC") +
  labs(x = "Chromosome") +
  theme_classic()+
  labs(y=expression(-Log10(q-value))) +
  scale_shape_manual(values=c(15,21,16,17))+
  scale_color_manual(values=c("grey", "dark grey"), guide =FALSE)+
  guides(shape = guide_legend(override.aes = list(color = "grey", size = 5, linetype =0) ) )

print(arc.STR)

ggsave(
  filename = "ARC_manhattan.png",
  plot     = arc.STR,
  width    = 10,      # in inches
  height   = 5,
  dpi      = 300,
  bg       = "transparent"  # or "white"
)
```

## NEH - Outlier Manhattan Plot
```{r}
neh_con_filter <- read.table("ToFilter.Con.Loci.NEH.txt", header = TRUE)
```

```{r, warning =FALSE, eval = FALSE}
neh.h <- subset(total.sig.NEH, Sig.STR == TRUE)
neh.c <- subset(NEH.pv, QCON < 0.1)


NEH.pp <- NEH.pv %>%
  mutate(
    Group = case_when(
      #SNP %in% singleton$SNP ~ "Singleton",
      SNP %in% neh.c$SNP ~ "Control Treatment Filtered",
      SNP %in% allcon.NEH$SNP ~ "Control Treatment Filtered",
      SNP %in% neh_con_filter$SNP ~ "Control Treatment Filtered",
      TRUE ~ "Shared SNP"
    )
  )

man <-ggman(NEH.pp, pvalue="QSTR", chr="CHR", snp="SNP",bp="BP", relative.positions = TRUE , sigLine = NA,
            pointSize=1.5,title = "", alpha = 0.5)

dfm <- man[[1]]

dfm$chrom_alt <- factor(dfm$chrom_alt, levels=c(0,1))

dfmo <- semi_join(dfm, neh.h, by = "SNP")
dfmo$Group <- "Outlier"

dfmsplit <- split(dfm, dfm$chrom)
xbreaks <- sapply(dfmsplit,function(x) {
    midpoint <- length(x$index)/2
    if(midpoint <1) midpoint <- 1
    return(x$index[midpoint])
})
neh.STR <-ggplot(dfm, aes(x= index, y=marker, colour = as.factor(chrom_alt)))+
  geom_point(aes(alpha=marker, shape=Group, size = marker))+  
  geom_point(data=dfmo, fill=cbPaletteSmall3[2], aes(x=index, y=marker, shape=Group, size=marker),  color = cbPaletteSmall3[2])+
  scale_x_continuous(breaks = xbreaks, labels = names(xbreaks),expand = c(0,0),limits = c(0,max(dfm$index)+10)) +
  guides(colour = FALSE,alpha=FALSE, fill = FALSE, size = FALSE) +
  #scale_y_continuous(expand = c(0,0),limits=c(0,max(dfm$marker+1,na.rm=TRUE)))+
  ylim(0,60) +
  scale_size_continuous(range = c(0.25,3)) +
  scale_alpha_continuous(range=c(0.05,0.8), limits= c(0,50))+
  ggtitle("NEH") +
  labs(x = "Chromosome") +
  theme_classic()+
  labs(y=expression(-Log10(q-value))) +
  scale_shape_manual(values=c(15,21,16,17))+
  scale_color_manual(values=c("grey", "dark grey"), guide =FALSE)+
  guides(shape = guide_legend(override.aes = list(color = "grey", size = 5, linetype =0) ) )

print(neh.STR)

ggsave(
  filename = "NEH_manhattan.png",
  plot     = neh.STR,
  width    = 10,      # in inches
  height   = 5,
  dpi      = 300,
  bg       = "transparent"  # or "white"
)
```
## MV - Outlier Manhattan Plot
```{r}
mv_con_filter <- read.table("ToFilter.Con.Loci.MV.txt", header = TRUE)
```

```{r, warning =FALSE, eval = FALSE}
mv.h <- subset(total.sig.MV, Sig.STR == TRUE)
mv.c <- subset(MV.pv, QCON < 0.1)


MV.pp <- MV.pv %>%
  mutate(
    Group = case_when(
      #SNP %in% singleton$SNP ~ "Singleton",
      SNP %in% mv.c$SNP ~ "Control Treatment Filtered",
      SNP %in% allcon.MV$SNP ~ "Control Treatment Filtered",
      SNP %in% mv_con_filter$SNP ~ "Control Treatment Filtered",
      TRUE ~ "Shared SNP"
    )
  )

man <-ggman(MV.pp, pvalue="QSTR", chr="CHR", snp="SNP",bp="BP", relative.positions = TRUE , sigLine = NA,
            pointSize=1.5,title = "", alpha = 0.5)

dfm <- man[[1]]

dfm$chrom_alt <- factor(dfm$chrom_alt, levels=c(0,1))

dfmo <- semi_join(dfm, mv.h, by = "SNP")
dfmo$Group <- "Outlier"

dfmsplit <- split(dfm, dfm$chrom)
xbreaks <- sapply(dfmsplit,function(x) {
    midpoint <- length(x$index)/2
    if(midpoint <1) midpoint <- 1
    return(x$index[midpoint])
})
mv.STR <-ggplot(dfm, aes(x= index, y=marker, colour = as.factor(chrom_alt)))+
  geom_point(aes(alpha=marker, shape=Group, size = marker))+  
  geom_point(data=dfmo, fill=cbPaletteSmall3[3], aes(x=index, y=marker, shape=Group, size=marker),  color = cbPaletteSmall3[3])+
  scale_x_continuous(breaks = xbreaks, labels = names(xbreaks),expand = c(0,0),limits = c(0,max(dfm$index)+10)) +
  guides(colour = FALSE,alpha=FALSE, fill = FALSE, size = FALSE) +
  scale_y_continuous(expand = c(0,0),limits=c(0,max(dfm$marker+1,na.rm=TRUE)))+
  scale_size_continuous(range = c(0.25,3)) +
  scale_alpha_continuous(range=c(0.05,0.8), limits= c(0,50))+
  ggtitle("MV") +
  labs(x = "Chromosome") +
  theme_classic()+
  labs(y=expression(-Log10(q-value))) +
  scale_shape_manual(values=c(15,21,16,17))+
  scale_color_manual(values=c("grey", "dark grey"), guide =FALSE)+
  guides(shape = guide_legend(override.aes = list(color = "grey", size = 5, linetype =0) ) )

print(mv.STR)

ggsave(
  filename = "MV_manhattan.png",
  plot     = mv.STR,
  width    = 10,      # in inches
  height   = 5,
  dpi      = 300,
  bg       = "transparent"  # or "white"
)
```












# Overlap in outlier loci
```{r}
library(dplyr)

arc <- read.table("Total.Significant.Loci.ARC", header = TRUE, sep = "\t", stringsAsFactors = FALSE)
neh <- read.table("Total.Significant.Loci.NEH", header = TRUE, sep = "\t", stringsAsFactors = FALSE)
mv  <- read.table("Total.Significant.Loci.MV",  header = TRUE, sep = "\t", stringsAsFactors = FALSE)

arc_sig <- unique(arc$SNP)
neh_sig <- unique(neh$SNP)
mv_sig  <- unique(mv$SNP)
```

```{r}
# All three populations
shared_all <- Reduce(intersect, list(arc_sig, neh_sig, mv_sig))
shared_all

# Pairwise overlaps
shared_arc_neh <- intersect(arc_sig, neh_sig)
shared_arc_mv  <- intersect(arc_sig, mv_sig)
shared_neh_mv  <- intersect(neh_sig, mv_sig)

# Counts
length(shared_all)
length(shared_arc_neh)
length(shared_arc_mv)
length(shared_neh_mv)

# Example: loci shared by ARC and NEH but not MV
arc_neh_only <- setdiff(shared_arc_neh, mv_sig)
length(arc_neh_only)

```

summary table
```{r}
overlap_summary <- data.frame(
  comparison = c("ARC_NEH", "ARC_MV", "NEH_MV", "ALL_THREE"),
  n_loci = c(
    length(shared_arc_neh),
    length(shared_arc_mv),
    length(shared_neh_mv),
    length(shared_all)
  )
)

overlap_summary
```

### Outlier Venn Diagram
```{r}
library(VennDiagram)
library(grid)

arc <- read.table("Total.Significant.Loci.ARC", header = TRUE, sep = "\t", stringsAsFactors = FALSE)
neh <- read.table("Total.Significant.Loci.NEH", header = TRUE, sep = "\t", stringsAsFactors = FALSE)
mv  <- read.table("Total.Significant.Loci.MV",  header = TRUE, sep = "\t", stringsAsFactors = FALSE)

arc_sig <- unique(arc$SNP)
neh_sig <- unique(neh$SNP)
mv_sig  <- unique(mv$SNP)

venn.plot.outlier <- venn.diagram(
  x = list(
    ARC = arc_sig,
    NEH = neh_sig,
    MV  = mv_sig
  ),
  filename = NULL,
  fill = c("#3F8F7A", "#D8848A", "#5F7BAA"),
  alpha = 0.5,
  cex = 1.5,
  cat.cex = 1.3,
  fontface = "plain",
  cat.fontface = "bold",
  cat.pos = c(315, 45, 0),
  cat.dist = c(0.05, 0.05, 0.05),
  margin = 0.02
)

grid.newpage()
grid.draw(venn.plot.outlier)

png("allpop_outliers_Venn.png", width = 2000, height = 2000, res = 300,  bg = "transparent")
grid.newpage()
grid.draw(venn.plot.outlier)
grid.text(
  "Shared candidate loci across populations",
  x = 0.5, y = 0.97,
  gp = gpar(fontsize = 18, fontface = "bold"))
dev.off()
```




# Outlier PCA 
PCAs on outliers from different populations

## ARC
```{bash, eval = FALSE}
# Activate the CASE conda environment
source activate CADO

mawk '!/CHR/' Total.Significant.Loci.ARC | cut -f2,3 | sort | uniq > TSL.pca.filter.ARC

mawk 'NR==FNR{a[$1,$2]; next} ($1,$2) in a' TSL.pca.filter.ARC ARC.dp25.cov.sync | cut --complement -f13- > ARC.outlier.pca.sync
```

```{r}
# Read sync file of significant outlier loci for outlier PCA
ARC.input.all.outlier <- read.sync(file="ARC.outlier.pca.sync", gen=seq(1,9), repl=seq(1,9), polarization="rising")
ARC.input.all.outlier.af <- af(ARC.input.all.outlier,gen=seq(1,9), repl=seq(1,9))

ARC.labels <- c(
  "ARC2_CB1","ARC2_CB2","ARC2_CB3","ARC2_SB1","ARC2_SB2","ARC2_SB3","ARC2_T01","ARC2_T02","ARC2_T03"
)
length(ARC.labels)  # should be 9

pop.ids <- colnames(ARC.input.all.outlier.af)  # "F1.R1.freq", ..., "F63.R63.freq"

meta <- data.frame(
  sample_id   = pop.ids,
  bio_label   = ARC.labels,
  stringsAsFactors = FALSE
)

# Derive treatment from the biological label
meta$treatment <- ifelse(grepl("_CB", meta$bio_label), "control",
                  ifelse(grepl("_SB", meta$bio_label), "stress", "initial"))
# can also parse by trial  
meta$trial <- sub("_.*", "", meta$bio_label)
# and parse by population
meta$pop <- ifelse(grepl("^ARC", meta$bio_label), "ARC",
                   ifelse(grepl("^NEH",  meta$bio_label),  "NEH",
                          ifelse(grepl("^MV", meta$bio_label), "MV", NA)))
View(meta)

#check that column names in input.all.rand.af and sample_id column in meta table are the same
all.equal(colnames(ARC.input.all.outlier.af), meta$sample_id)
# should be TRUE

# assign the columns of input.all.outlier.af with their proper sample names instead of the sample_id created by inputting the table to poolseq
colnames(ARC.input.all.outlier.af) <- meta$bio_label
```

```{r}
# Perform PCA
af_t <- t(ARC.input.all.outlier.af)

# Run PCA on transposed allele frequency matrix (samples as rows, loci as columns)
pca_res <- prcomp(af_t, center = TRUE, scale. = TRUE)

# Get scores for plotting
pca_df <- as.data.frame(pca_res$x)
#pca_df$locus <- rownames(input.all.outlier.af)  # keep locus IDs if you have them

# Calculate percent variance explained by each PC
var_explained <- pca_res$sdev^2 / sum(pca_res$sdev^2) * 100

pca_df$Treatment <- meta$treatment
pca_df$Trial <- meta$trial

outpca_arc <- ggplot(pca_df, aes(x = PC1, y = PC2, color = Treatment, fill=Treatment)) +
  geom_point(aes(alpha=0.2), size =7,shape=21, col="black")+  
  scale_fill_manual(values=treatPalette,name="Treatment")+ 
guides(alpha=FALSE) +theme(axis.title.x = element_text(size = 24),axis.title.y = element_text(size = 24)) +
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 1)
    ) +
  labs(
    x = paste0("PC1 (", round(var_explained[1], 1), "%)"),
    y = paste0("PC2 (", round(var_explained[2], 1), "%)"),
    title = "PCA of Outlier Loci (ARC)"
  )

ggsave(filename = "PCA/outlier_PCA_ARC.png", plot = outpca_arc, width = 6, height = 5, units = "in", dpi = 300)
```

## NEH
```{bash, eval = FALSE}
# Activate the CASE conda environment
source activate CADO

mawk '!/CHR/' Total.Significant.Loci.NEH | cut -f2,3 | sort | uniq > TSL.pca.filter.NEH

mawk 'NR==FNR{a[$1,$2]; next} ($1,$2) in a' TSL.pca.filter.NEH NEH.dp25.cov.sync | cut --complement -f31- > NEH.outlier.pca.sync
```

```{r}
# Read sync file of significant outlier loci for outlier PCA
NEH.input.all.outlier <- read.sync(file="NEH.outlier.pca.sync", gen=seq(1,27), repl=seq(1,27), polarization="rising")
NEH.input.all.outlier.af <- af(NEH.input.all.outlier,gen=seq(1,27), repl=seq(1,27))

NEH.labels <- c(
  "NEH1_CB1","NEH1_CB2","NEH1_CB3","NEH1_SB1","NEH1_SB2","NEH1_SB3","NEH1_T01","NEH1_T02","NEH1_T03",
  "NEH2_CB1","NEH2_CB2","NEH2_CB3","NEH2_SB1","NEH2_SB2","NEH2_SB3","NEH2_T01","NEH2_T02","NEH2_T03",
  "NEH3_CB1","NEH3_CB2","NEH3_CB3","NEH3_SB1","NEH3_SB2","NEH3_SB3","NEH3_T01","NEH3_T02","NEH3_T03"
)
length(NEH.labels)  # should be 27

pop.ids <- colnames(NEH.input.all.outlier.af)  # "F1.R1.freq", ..., "F63.R63.freq"

meta <- data.frame(
  sample_id   = pop.ids,
  bio_label   = NEH.labels,
  stringsAsFactors = FALSE
)

# Derive treatment from the biological label
meta$treatment <- ifelse(grepl("_CB", meta$bio_label), "control",
                  ifelse(grepl("_SB", meta$bio_label), "stress", "initial"))
# can also parse by trial  
meta$trial <- sub("_.*", "", meta$bio_label)
# and parse by population
meta$pop <- ifelse(grepl("^ARC", meta$bio_label), "ARC",
                   ifelse(grepl("^NEH",  meta$bio_label),  "NEH",
                          ifelse(grepl("^MV", meta$bio_label), "MV", NA)))
View(meta)

#check that column names in input.all.rand.af and sample_id column in meta table are the same
all.equal(colnames(NEH.input.all.outlier.af), meta$sample_id)
# should be TRUE

# assign the columns of input.all.outlier.af with their proper sample names instead of the sample_id created by inputting the table to poolseq
colnames(NEH.input.all.outlier.af) <- meta$bio_label
```

```{r}
# Perform PCA
af_t <- t(NEH.input.all.outlier.af)

# Run PCA on transposed allele frequency matrix (samples as rows, loci as columns)
pca_res <- prcomp(af_t, center = TRUE, scale. = TRUE)

# Get scores for plotting
pca_df <- as.data.frame(pca_res$x)
#pca_df$locus <- rownames(input.all.outlier.af)  # keep locus IDs if you have them

# Calculate percent variance explained by each PC
var_explained <- pca_res$sdev^2 / sum(pca_res$sdev^2) * 100

pca_df$Treatment <- meta$treatment
pca_df$Trial <- meta$trial

outpca_neh <- ggplot(pca_df, aes(x = PC1, y = PC2, color = Treatment, fill=Treatment)) +
  geom_point(aes(alpha=0.2), size =7,shape=21, col="black")+  
  scale_fill_manual(values=treatPalette,name="Treatment")+ 
  annotate("text", x = -30, y = 3, label = "NEH1", color = "black", size = 5) +
  annotate("text", x = -3, y = -20, label = "NEH3", color = "black", size = 5) +
  annotate("text", x = 22, y = 6, label = "NEH2", color = "black", size = 5) +
guides(alpha=FALSE) +theme(axis.title.x = element_text(size = 24),axis.title.y = element_text(size = 24)) +
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 1)
    ) +
  labs(
    x = paste0("PC1 (", round(var_explained[1], 1), "%)"),
    y = paste0("PC2 (", round(var_explained[2], 1), "%)"),
    title = "PCA of Outlier Loci (NEH)"
  )
outpca_neh
ggsave(filename = "PCA/outlier_PCA_NEH.png", plot = outpca_neh, width = 6, height = 5, units = "in", dpi = 300)
```

## MV
```{bash, eval = FALSE}
# Activate the CASE conda environment
source activate CADO

mawk '!/CHR/' Total.Significant.Loci.MV | cut -f2,3 | sort | uniq > TSL.pca.filter.MV

mawk 'NR==FNR{a[$1,$2]; next} ($1,$2) in a' TSL.pca.filter.MV MV.dp25.cov.sync | cut --complement -f31- > MV.outlier.pca.sync
```

```{r}
# Read sync file of significant outlier loci for outlier PCA
MV.input.all.outlier <- read.sync(file="MV.outlier.pca.sync", gen=seq(1,27), repl=seq(1,27), polarization="rising")
MV.input.all.outlier.af <- af(MV.input.all.outlier,gen=seq(1,27), repl=seq(1,27))

MV.labels <- c(
  "MV1_CB1","MV1_CB2","MV1_CB3","MV1_SB1","MV1_SB2","MV1_SB3","MV1_T01","MV1_T02","MV1_T03",
  "MV2_CB1","MV2_CB2","MV2_CB3","MV2_SB1","MV2_SB2","MV2_SB3","MV2_T01","MV2_T02","MV2_T03",
  "MV3_CB1","MV3_CB2","MV3_CB3","MV3_SB1","MV3_SB2","MV3_SB3","MV3_T01","MV3_T02","MV3_T03"
)
length(MV.labels)  # should be 27

pop.ids <- colnames(MV.input.all.outlier.af)  # "F1.R1.freq", ..., "F63.R63.freq"

meta <- data.frame(
  sample_id   = pop.ids,
  bio_label   = MV.labels,
  stringsAsFactors = FALSE
)

# Derive treatment from the biological label
meta$treatment <- ifelse(grepl("_CB", meta$bio_label), "control",
                  ifelse(grepl("_SB", meta$bio_label), "stress", "initial"))
# can also parse by trial  
meta$trial <- sub("_.*", "", meta$bio_label)
# and parse by population
meta$pop <- ifelse(grepl("^ARC", meta$bio_label), "ARC",
                   ifelse(grepl("^MV",  meta$bio_label),  "MV",
                          ifelse(grepl("^NEH", meta$bio_label), "NEH", NA)))
View(meta)

#check that column names in input.all.rand.af and sample_id column in meta table are the same
all.equal(colnames(MV.input.all.outlier.af), meta$sample_id)
# should be TRUE

# assign the columns of input.all.outlier.af with their proper sample names instead of the sample_id created by inputting the table to poolseq
colnames(MV.input.all.outlier.af) <- meta$bio_label
```

```{r}
# Perform PCA
af_t <- t(MV.input.all.outlier.af)

# Run PCA on transposed allele frequency matrix (samples as rows, loci as columns)
pca_res <- prcomp(af_t, center = TRUE, scale. = TRUE)

# Get scores for plotting
pca_df <- as.data.frame(pca_res$x)
#pca_df$locus <- rownames(input.all.outlier.af)  # keep locus IDs if you have them

# Calculate percent variance explained by each PC
var_explained <- pca_res$sdev^2 / sum(pca_res$sdev^2) * 100

pca_df$Treatment <- meta$treatment
pca_df$Trial <- meta$trial

outpca_mv <- ggplot(pca_df, aes(x = PC1, y = PC2, color = Treatment, fill=Treatment)) +
  geom_point(aes(alpha=0.2), size =7,shape=21, col="black")+  
  scale_fill_manual(values=treatPalette,name="Treatment")+ 
  annotate("text", x = -25, y = -17, label = "MV1", color = "black", size = 5) +
  annotate("text", x = -17, y = 30, label = "MV2", color = "black", size = 5) +
  annotate("text", x = 35, y = 13, label = "MV3", color = "black", size = 5) +
guides(alpha=FALSE) +theme(axis.title.x = element_text(size = 24),axis.title.y = element_text(size = 24)) +
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 1)
    ) +
  labs(
    x = paste0("PC1 (", round(var_explained[1], 1), "%)"),
    y = paste0("PC2 (", round(var_explained[2], 1), "%)"),
    title = "PCA of Outlier Loci (MV)"
  )
outpca_mv

ggsave(filename = "PCA/outlier_PCA_MV.png", plot = outpca_mv, width = 6, height = 5, units = "in", dpi = 300)
```


## Composite Outlier PCA with 3 pops
```{r}
outpca_design <- "AAABBBCCC"

allpop_outpca <- (outpca_arc + outpca_neh + outpca_mv) +
  plot_layout(design = outpca_design, guides = "collect") +
  plot_annotation(tag_levels = "a", tag_suffix = ")") 
print(allpop_outpca)

# Export to PNG at a specific size
ggsave("PCA/outlier_pca_allpopcomposite.png", plot = allpop_outpca, width = 15, height =5, units = "in", dpi = 300)
```



# Outlier Fst tables
## Pairwise Fst

```{r}
library(poolfstat)
```

Construct outlier sync files for each population!
```{bash, eval=FALSE}

#ARC 
#pull CHROM & BP from outlier loci file and sorts them into position file
# keeps header from sync file and prints only rows whose "CHROM"_"POS" matches the positions file 
awk 'BEGIN{FS=OFS="\t"} NR>1{print $2,$3}' Total.Significant.Loci.ARC | sort -u > ARC.outlier.positions.txt && awk 'BEGIN{FS=OFS="\t"} NR==FNR{if(FNR>1) keep[$1"_"$2]=1; next} FNR==1 || ($1"_"$2 in keep)' ARC.outlier.positions.txt ARC.dp25.cov.sync > ARC.outliers.only.sync
# NEH
awk 'BEGIN{FS=OFS="\t"} NR>1{print $2,$3}' Total.Significant.Loci.NEH | sort -u > NEH.outlier.positions.txt && awk 'BEGIN{FS=OFS="\t"} NR==FNR{if(FNR>1) keep[$1"_"$2]=1; next} FNR==1 || ($1"_"$2 in keep)' NEH.outlier.positions.txt NEH.dp25.cov.sync > NEH.outliers.only.sync
#MV
awk 'BEGIN{FS=OFS="\t"} NR>1{print $2,$3}' Total.Significant.Loci.MV | sort -u > MV.outlier.positions.txt && awk 'BEGIN{FS=OFS="\t"} NR==FNR{if(FNR>1) keep[$1"_"$2]=1; next} FNR==1 || ($1"_"$2 in keep)' MV.outlier.positions.txt MV.dp25.cov.sync > MV.outliers.only.sync

#check that these match the correct # of outliers
wc -l ARC.outliers.only.sync 
wc -l NEH.outliers.only.sync
wc -l MV.outliers.only.sync

#remove extra lines with coverage stats
awk 'BEGIN{FS=OFS="\t"} {NF-=3; print}' ARC.outliers.only.sync > ARC.outliers.only.nocov.sync
awk 'BEGIN{FS=OFS="\t"} {NF-=3; print}' NEH.outliers.only.sync > NEH.outliers.only.nocov.sync
awk 'BEGIN{FS=OFS="\t"} {NF-=3; print}' MV.outliers.only.sync > MV.outliers.only.nocov.sync

# remove header line 
mawk '!/CHR/' ARC.outliers.only.nocov.sync > ARC.total.outlier.input
mawk '!/CHR/' NEH.outliers.only.nocov.sync > NEH.total.outlier.input
mawk '!/CHR/' MV.outliers.only.nocov.sync > MV.total.outlier.input
```


## ARC
```{r}
##Using sync file with outlier for each population 
out_ARC_samples <- c("ARC2_CB1","ARC2_CB2","ARC2_CB3","ARC2_SB1","ARC2_SB2","ARC2_SB3","ARC2_T01","ARC2_T02","ARC2_T03")

out_ARCdstbn <- popsync2pooldata("ARC.total.outlier.input", poolsizes = c(40000, 40000, 40000, 40000,
                                                                          40000, 40000, 40000, 40000, 40000), 
                                 poolnames = out_ARC_samples, min.rc = 1, min.cov.per.pool = 1, 
                                 max.cov.per.pool = 50000, min.maf = 0.1, noindel = TRUE, nlines.per.readblock = 1000)

#Compute Fst for every SNP
out_ARCdstbn_SNPFst <- computeFST(out_ARCdstbn, method = "Anova")

#Compute pairwise Fst matrix
out_ARCdstbn_Fst <- compute.pairwiseFST(out_ARCdstbn, method = "Anova", min.cov.per.pool = 1, max.cov.per.pool = 50000, min.maf = 0.001, output.snp.values = FALSE)

out_ARCdstbn_PairwiseFSTMatrix <- as.matrix(as.data.frame(out_ARCdstbn_Fst@PairwiseFSTmatrix))

ARC_Fst <- out_ARCdstbn_PairwiseFSTMatrix[c("ARC2_CB1","ARC2_CB2","ARC2_CB3","ARC2_SB1","ARC2_SB2","ARC2_SB3","ARC2_T01","ARC2_T02","ARC2_T03"), c("ARC2_CB1","ARC2_CB2","ARC2_CB3","ARC2_SB1","ARC2_SB2","ARC2_SB3","ARC2_T01","ARC2_T02","ARC2_T03")]
write.csv(ARC_Fst, "ARC_Fst_Outliers.csv")
```

## NEH
```{r}
##Using sync file with outlier for each population 
out_NEH_samples <- c("NEH1_CB1","NEH1_CB2","NEH1_CB3","NEH1_SB1","NEH1_SB2","NEH1_SB3","NEH1_T01","NEH1_T02","NEH1_T03",
                     "NEH2_CB1","NEH2_CB2","NEH2_CB3","NEH2_SB1","NEH2_SB2","NEH2_SB3","NEH2_T01","NEH2_T02","NEH2_T03",
                     "NEH3_CB1","NEH3_CB2","NEH3_CB3","NEH3_SB1","NEH3_SB2","NEH3_SB3","NEH3_T01","NEH3_T02","NEH3_T03")

#update poolsizes to reflect 27 samples
out_NEHdstbn <- popsync2pooldata("NEH.total.outlier.input", poolsizes = c(20000,20000, 20000, 20000, 20000, 20000,40000, 40000, 40000, 
                                                                          20000,20000, 20000, 20000, 20000, 20000,40000, 40000, 40000, 
                                                                          40000, 40000, 40000, 40000,40000, 40000, 40000, 40000, 40000), 
                                 poolnames = out_NEH_samples, min.rc = 1, min.cov.per.pool = 1, 
                                 max.cov.per.pool = 50000, min.maf = 0.1, noindel = TRUE, nlines.per.readblock = 1000)

#Compute Fst for every SNP
out_NEHdstbn_SNPFst <- computeFST(out_NEHdstbn, method = "Anova")

#Compute pairwise Fst matrix
out_NEHdstbn_Fst <- compute.pairwiseFST(out_NEHdstbn, method = "Anova", min.cov.per.pool = 1, max.cov.per.pool = 50000, min.maf = 0.001, output.snp.values = FALSE)

out_NEHdstbn_PairwiseFSTMatrix <- as.matrix(as.data.frame(out_NEHdstbn_Fst@PairwiseFSTmatrix))

NEH_Fst <- out_NEHdstbn_PairwiseFSTMatrix[c("NEH1_CB1","NEH1_CB2","NEH1_CB3","NEH1_SB1","NEH1_SB2","NEH1_SB3","NEH1_T01","NEH1_T02","NEH1_T03",
                     "NEH2_CB1","NEH2_CB2","NEH2_CB3","NEH2_SB1","NEH2_SB2","NEH2_SB3","NEH2_T01","NEH2_T02","NEH2_T03",
                     "NEH3_CB1","NEH3_CB2","NEH3_CB3","NEH3_SB1","NEH3_SB2","NEH3_SB3","NEH3_T01","NEH3_T02","NEH3_T03"), c("NEH1_CB1","NEH1_CB2","NEH1_CB3","NEH1_SB1","NEH1_SB2","NEH1_SB3","NEH1_T01","NEH1_T02","NEH1_T03",
                     "NEH2_CB1","NEH2_CB2","NEH2_CB3","NEH2_SB1","NEH2_SB2","NEH2_SB3","NEH2_T01","NEH2_T02","NEH2_T03",
                     "NEH3_CB1","NEH3_CB2","NEH3_CB3","NEH3_SB1","NEH3_SB2","NEH3_SB3","NEH3_T01","NEH3_T02","NEH3_T03")]
write.csv(NEH_Fst, "NEH_Fst_Outliers.csv")
```

## MV
```{r}
##Using sync file with outlier for each population 
out_MV_samples <- c("MV1_CB1","MV1_CB2","MV1_CB3","MV1_SB1","MV1_SB2","MV1_SB3","MV1_T01","MV1_T02","MV1_T03",
                     "MV2_CB1","MV2_CB2","MV2_CB3","MV2_SB1","MV2_SB2","MV2_SB3","MV2_T01","MV2_T02","MV2_T03",
                     "MV3_CB1","MV3_CB2","MV3_CB3","MV3_SB1","MV3_SB2","MV3_SB3","MV3_T01","MV3_T02","MV3_T03")

#update poolsizes to reflect 27 samples
out_MVdstbn <- popsync2pooldata("MV.total.outlier.input", poolsizes = c(40000, 40000, 40000, 40000,40000, 40000, 40000, 40000, 40000, 
                                                                          40000, 40000, 40000, 40000,40000, 40000, 40000, 40000, 40000, 
                                                                          40000, 40000, 40000, 40000,40000, 40000, 40000, 40000, 40000), 
                                 poolnames = out_MV_samples, min.rc = 1, min.cov.per.pool = 1, 
                                 max.cov.per.pool = 50000, min.maf = 0.1, noindel = TRUE, nlines.per.readblock = 1000)

#Compute Fst for every SNP
out_MVdstbn_SNPFst <- computeFST(out_MVdstbn, method = "Anova")

#Compute pairwise Fst matrix
out_MVdstbn_Fst <- compute.pairwiseFST(out_MVdstbn, method = "Anova", min.cov.per.pool = 1, max.cov.per.pool = 50000, min.maf = 0.001, output.snp.values = FALSE)

out_MVdstbn_PairwiseFSTMatrix <- as.matrix(as.data.frame(out_MVdstbn_Fst@PairwiseFSTmatrix))

MV_Fst <- out_MVdstbn_PairwiseFSTMatrix[c("MV1_CB1","MV1_CB2","MV1_CB3","MV1_SB1","MV1_SB2","MV1_SB3","MV1_T01","MV1_T02","MV1_T03",
                     "MV2_CB1","MV2_CB2","MV2_CB3","MV2_SB1","MV2_SB2","MV2_SB3","MV2_T01","MV2_T02","MV2_T03",
                     "MV3_CB1","MV3_CB2","MV3_CB3","MV3_SB1","MV3_SB2","MV3_SB3","MV3_T01","MV3_T02","MV3_T03"), c("MV1_CB1","MV1_CB2","MV1_CB3","MV1_SB1","MV1_SB2","MV1_SB3","MV1_T01","MV1_T02","MV1_T03",
                     "MV2_CB1","MV2_CB2","MV2_CB3","MV2_SB1","MV2_SB2","MV2_SB3","MV2_T01","MV2_T02","MV2_T03",
                     "MV3_CB1","MV3_CB2","MV3_CB3","MV3_SB1","MV3_SB2","MV3_SB3","MV3_T01","MV3_T02","MV3_T03")]
write.csv(MV_Fst, "MV_Fst_Outliers.csv")
```



# Neutral loci
### Prep Neutral loci files
Make neutral loci sync files for each population
```{bash, eval=FALSE}
# ARC
awk 'BEGIN{FS=OFS="\t"} { NF-=3; print }' ARC.dp25.cov.sync > ARC.dp25.cov.noStats.sync
awk 'NR==FNR { if (FNR>1) bad[$1]=1; next }
     FNR==1 { print; next }
    { key=$1"_"$2; if (!(key in bad)) print }' \
Total.Significant.Loci.ARC ARC.dp25.cov.noStats.sync > ARC.neutral.only.sync

#NEH
awk 'BEGIN{FS=OFS="\t"} { NF-=3; print }' NEH.dp25.cov.sync > NEH.dp25.cov.noStats.sync
awk 'NR==FNR { if (FNR>1) bad[$1]=1; next }
     FNR==1 { print; next }
    { key=$1"_"$2; if (!(key in bad)) print }' \
Total.Significant.Loci.NEH NEH.dp25.cov.noStats.sync > NEH.neutral.only.sync

#MV
awk 'BEGIN{FS=OFS="\t"} { NF-=3; print }' MV.dp25.cov.sync > MV.dp25.cov.noStats.sync
awk 'NR==FNR { if (FNR>1) bad[$1]=1; next }
     FNR==1 { print; next }
    { key=$1"_"$2; if (!(key in bad)) print }' \
Total.Significant.Loci.MV MV.dp25.cov.noStats.sync > MV.neutral.only.sync
```

```{bash}
wc -l ARC.neutral.only.sync
wc -l NEH.neutral.only.sync
wc -l MV.neutral.only.sync
```

## Neutral Loci Fst
```{bash}
#remove extra lines with coverage stats
#awk 'BEGIN{FS=OFS="\t"} {NF-=3; print}' ARC.neutral.only.sync > ARC.neutral.only.nocov.sync
#awk 'BEGIN{FS=OFS="\t"} {NF-=3; print}' NEH.neutral.only.sync > NEH.neutral.only.nocov.sync
#awk 'BEGIN{FS=OFS="\t"} {NF-=3; print}' MV.neutral.only.sync > MV.neutral.only.nocov.sync

# remove header line 
mawk '!/CHR/' ARC.neutral.only.sync > ARC.total.neutral.input
mawk '!/CHR/' NEH.neutral.only.sync > NEH.total.neutral.input
mawk '!/CHR/' MV.neutral.only.sync > MV.total.neutral.input
```


### ARC
```{r}
##Using sync file with neutral for each population 
neut_ARC_samples <- c("ARC2_CB1","ARC2_CB2","ARC2_CB3","ARC2_SB1","ARC2_SB2","ARC2_SB3","ARC2_T01","ARC2_T02","ARC2_T03")

neut_ARCdstbn <- popsync2pooldata("ARC.total.neutral.input", poolsizes = c(40000, 40000, 40000, 40000,
                                                                          40000, 40000, 40000, 40000, 40000), 
                                 poolnames = neut_ARC_samples, min.rc = 1, min.cov.per.pool = 1, 
                                 max.cov.per.pool = 50000, min.maf = 0.1, noindel = TRUE, nlines.per.readblock = 1000)

#Compute Fst for every SNP
neut_ARCdstbn_SNPFst <- computeFST(neut_ARCdstbn, method = "Anova")

#Compute pairwise Fst matrix
neut_ARCdstbn_Fst <- compute.pairwiseFST(neut_ARCdstbn, method = "Anova", min.cov.per.pool = 1, max.cov.per.pool = 50000, min.maf = 0.001, output.snp.values = FALSE)

neut_ARCdstbn_PairwiseFSTMatrix <- as.matrix(as.data.frame(neut_ARCdstbn_Fst@PairwiseFSTmatrix))

ARC_Fst <- neut_ARCdstbn_PairwiseFSTMatrix[c("ARC2_CB1","ARC2_CB2","ARC2_CB3","ARC2_SB1","ARC2_SB2","ARC2_SB3","ARC2_T01","ARC2_T02","ARC2_T03"), c("ARC2_CB1","ARC2_CB2","ARC2_CB3","ARC2_SB1","ARC2_SB2","ARC2_SB3","ARC2_T01","ARC2_T02","ARC2_T03")]
write.csv(ARC_Fst, "ARC_Fst_Neutral.csv")
```

### NEH
```{r}
##Using sync file with neutral for each population 
neut_NEH_samples <- c("NEH1_CB1","NEH1_CB2","NEH1_CB3","NEH1_SB1","NEH1_SB2","NEH1_SB3","NEH1_T01","NEH1_T02","NEH1_T03",
                     "NEH2_CB1","NEH2_CB2","NEH2_CB3","NEH2_SB1","NEH2_SB2","NEH2_SB3","NEH2_T01","NEH2_T02","NEH2_T03",
                     "NEH3_CB1","NEH3_CB2","NEH3_CB3","NEH3_SB1","NEH3_SB2","NEH3_SB3","NEH3_T01","NEH3_T02","NEH3_T03")

#update poolsizes to reflect 27 samples
neut_NEHdstbn <- popsync2pooldata("NEH.total.neutral.input", poolsizes = c(20000,20000, 20000, 20000, 20000, 20000,40000, 40000, 40000, 
                                                                          20000,20000, 20000, 20000, 20000, 20000,40000, 40000, 40000, 
                                                                          40000, 40000, 40000, 40000,40000, 40000, 40000, 40000, 40000), 
                                 poolnames = neut_NEH_samples, min.rc = 1, min.cov.per.pool = 1, 
                                 max.cov.per.pool = 50000, min.maf = 0.1, noindel = TRUE, nlines.per.readblock = 1000)

#Compute Fst for every SNP
neut_NEHdstbn_SNPFst <- computeFST(neut_NEHdstbn, method = "Anova")

#Compute pairwise Fst matrix
neut_NEHdstbn_Fst <- compute.pairwiseFST(neut_NEHdstbn, method = "Anova", min.cov.per.pool = 1, max.cov.per.pool = 50000, min.maf = 0.001, output.snp.values = FALSE)

neut_NEHdstbn_PairwiseFSTMatrix <- as.matrix(as.data.frame(neut_NEHdstbn_Fst@PairwiseFSTmatrix))

NEH_Fst <- neut_NEHdstbn_PairwiseFSTMatrix[c("NEH1_CB1","NEH1_CB2","NEH1_CB3","NEH1_SB1","NEH1_SB2","NEH1_SB3","NEH1_T01","NEH1_T02","NEH1_T03",
                     "NEH2_CB1","NEH2_CB2","NEH2_CB3","NEH2_SB1","NEH2_SB2","NEH2_SB3","NEH2_T01","NEH2_T02","NEH2_T03",
                     "NEH3_CB1","NEH3_CB2","NEH3_CB3","NEH3_SB1","NEH3_SB2","NEH3_SB3","NEH3_T01","NEH3_T02","NEH3_T03"), c("NEH1_CB1","NEH1_CB2","NEH1_CB3","NEH1_SB1","NEH1_SB2","NEH1_SB3","NEH1_T01","NEH1_T02","NEH1_T03",
                     "NEH2_CB1","NEH2_CB2","NEH2_CB3","NEH2_SB1","NEH2_SB2","NEH2_SB3","NEH2_T01","NEH2_T02","NEH2_T03",
                     "NEH3_CB1","NEH3_CB2","NEH3_CB3","NEH3_SB1","NEH3_SB2","NEH3_SB3","NEH3_T01","NEH3_T02","NEH3_T03")]
write.csv(NEH_Fst, "NEH_Fst_Neutral.csv")
```

### MV
```{r}
##Using sync file with neutral for each population 
neut_MV_samples <- c("MV1_CB1","MV1_CB2","MV1_CB3","MV1_SB1","MV1_SB2","MV1_SB3","MV1_T01","MV1_T02","MV1_T03",
                     "MV2_CB1","MV2_CB2","MV2_CB3","MV2_SB1","MV2_SB2","MV2_SB3","MV2_T01","MV2_T02","MV2_T03",
                     "MV3_CB1","MV3_CB2","MV3_CB3","MV3_SB1","MV3_SB2","MV3_SB3","MV3_T01","MV3_T02","MV3_T03")

#update poolsizes to reflect 27 samples
neut_MVdstbn <- popsync2pooldata("MV.total.neutral.input", poolsizes = c(40000, 40000, 40000, 40000,40000, 40000, 40000, 40000, 40000, 
                                                                          40000, 40000, 40000, 40000,40000, 40000, 40000, 40000, 40000, 
                                                                          40000, 40000, 40000, 40000,40000, 40000, 40000, 40000, 40000), 
                                 poolnames = neut_MV_samples, min.rc = 1, min.cov.per.pool = 1, 
                                 max.cov.per.pool = 50000, min.maf = 0.1, noindel = TRUE, nlines.per.readblock = 1000)

#Compute Fst for every SNP
neut_MVdstbn_SNPFst <- computeFST(neut_MVdstbn, method = "Anova")

#Compute pairwise Fst matrix
neut_MVdstbn_Fst <- compute.pairwiseFST(neut_MVdstbn, method = "Anova", min.cov.per.pool = 1, max.cov.per.pool = 50000, min.maf = 0.001, output.snp.values = FALSE)

neut_MVdstbn_PairwiseFSTMatrix <- as.matrix(as.data.frame(neut_MVdstbn_Fst@PairwiseFSTmatrix))

MV_Fst <- neut_MVdstbn_PairwiseFSTMatrix[c("MV1_CB1","MV1_CB2","MV1_CB3","MV1_SB1","MV1_SB2","MV1_SB3","MV1_T01","MV1_T02","MV1_T03",
                     "MV2_CB1","MV2_CB2","MV2_CB3","MV2_SB1","MV2_SB2","MV2_SB3","MV2_T01","MV2_T02","MV2_T03",
                     "MV3_CB1","MV3_CB2","MV3_CB3","MV3_SB1","MV3_SB2","MV3_SB3","MV3_T01","MV3_T02","MV3_T03"), c("MV1_CB1","MV1_CB2","MV1_CB3","MV1_SB1","MV1_SB2","MV1_SB3","MV1_T01","MV1_T02","MV1_T03",
                     "MV2_CB1","MV2_CB2","MV2_CB3","MV2_SB1","MV2_SB2","MV2_SB3","MV2_T01","MV2_T02","MV2_T03",
                     "MV3_CB1","MV3_CB2","MV3_CB3","MV3_SB1","MV3_SB2","MV3_SB3","MV3_T01","MV3_T02","MV3_T03")]
write.csv(MV_Fst, "MV_Fst_Neutral.csv")
```


## Neutral Loci PCAs

### ARC
```{bash, eval = FALSE}
# Activate the CASE conda environment
source activate CADO

mawk '!/CHR/' ARC.neutral.only.sync | cut -f1,2 | sort | uniq > neutral.pca.filter.ARC

mawk 'NR==FNR{a[$1,$2]; next} ($1,$2) in a' neutral.pca.filter.ARC ARC.dp25.cov.sync | cut --complement -f13- > ARC.neutral.pca.sync
```

```{r}
# Read sync file of significant neutral loci for neutral PCA
ARC.input.all.neutral <- read.sync(file="ARC.neutral.pca.sync", gen=seq(1,9), repl=seq(1,9), polarization="rising")
ARC.input.all.neutral.af <- af(ARC.input.all.neutral,gen=seq(1,9), repl=seq(1,9))

ARC.labels <- c(
  "ARC2_CB1","ARC2_CB2","ARC2_CB3","ARC2_SB1","ARC2_SB2","ARC2_SB3","ARC2_T01","ARC2_T02","ARC2_T03"
)
length(ARC.labels)  # should be 9

pop.ids <- colnames(ARC.input.all.neutral.af)  # "F1.R1.freq", ..., "F63.R63.freq"

meta <- data.frame(
  sample_id   = pop.ids,
  bio_label   = ARC.labels,
  stringsAsFactors = FALSE
)

# Derive treatment from the biological label
meta$treatment <- ifelse(grepl("_CB", meta$bio_label), "control",
                  ifelse(grepl("_SB", meta$bio_label), "stress", "initial"))
# can also parse by trial  
meta$trial <- sub("_.*", "", meta$bio_label)
# and parse by population
meta$pop <- ifelse(grepl("^ARC", meta$bio_label), "ARC",
                   ifelse(grepl("^NEH",  meta$bio_label),  "NEH",
                          ifelse(grepl("^MV", meta$bio_label), "MV", NA)))
View(meta)

#check that column names in input.all.rand.af and sample_id column in meta table are the same
all.equal(colnames(ARC.input.all.neutral.af), meta$sample_id)
# should be TRUE

# assign the columns of input.all.neutral.af with their proper sample names instead of the sample_id created by inputting the table to poolseq
colnames(ARC.input.all.neutral.af) <- meta$bio_label
```

```{r}
# Perform PCA
af_t <- t(ARC.input.all.neutral.af)

# Run PCA on transposed allele frequency matrix (samples as rows, loci as columns)
pca_res <- prcomp(af_t, center = TRUE, scale. = TRUE)

# Get scores for plotting
pca_df <- as.data.frame(pca_res$x)
#pca_df$locus <- rownames(input.all.neutral.af)  # keep locus IDs if you have them

# Calculate percent variance explained by each PC
var_explained <- pca_res$sdev^2 / sum(pca_res$sdev^2) * 100

pca_df$Treatment <- meta$treatment
pca_df$Trial <- meta$trial

neutpca_arc <- ggplot(pca_df, aes(x = PC1, y = PC2, color = Treatment, fill=Treatment)) +
  geom_point(aes(alpha=0.2), size =7,shape=21, col="black")+  
  scale_fill_manual(values=treatPalette,name="Treatment")+ 
guides(alpha=FALSE) +theme(axis.title.x = element_text(size = 24),axis.title.y = element_text(size = 24)) +
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 1)
    ) +
  labs(
    x = paste0("PC1 (", round(var_explained[1], 1), "%)"),
    y = paste0("PC2 (", round(var_explained[2], 1), "%)"),
    title = "PCA of Neutral Loci (ARC)"
  )

ggsave(filename = "PCA/neutral_PCA_ARC.png", plot = neutpca_arc, width = 6, height = 5, units = "in", dpi = 300)
```

### NEH
```{bash, eval = FALSE}
# Activate the CASE conda environment
source activate CADO

mawk '!/CHR/' NEH.neutral.only.sync | cut -f1,2 | sort | uniq > neutral.pca.filter.NEH

mawk 'NR==FNR{a[$1,$2]; next} ($1,$2) in a' neutral.pca.filter.NEH NEH.dp25.cov.sync | cut --complement -f31- > NEH.neutral.pca.sync
```

```{r}
# Read sync file of significant neutral loci for neutral PCA
NEH.input.all.neutral <- read.sync(file="NEH.neutral.pca.sync", gen=seq(1,27), repl=seq(1,27), polarization="rising")
NEH.input.all.neutral.af <- af(NEH.input.all.neutral,gen=seq(1,27), repl=seq(1,27))

NEH.labels <- c(
  "NEH1_CB1","NEH1_CB2","NEH1_CB3","NEH1_SB1","NEH1_SB2","NEH1_SB3","NEH1_T01","NEH1_T02","NEH1_T03",
  "NEH2_CB1","NEH2_CB2","NEH2_CB3","NEH2_SB1","NEH2_SB2","NEH2_SB3","NEH2_T01","NEH2_T02","NEH2_T03",
  "NEH3_CB1","NEH3_CB2","NEH3_CB3","NEH3_SB1","NEH3_SB2","NEH3_SB3","NEH3_T01","NEH3_T02","NEH3_T03"
)
length(NEH.labels)  # should be 27

pop.ids <- colnames(NEH.input.all.neutral.af)  # "F1.R1.freq", ..., "F63.R63.freq"

meta <- data.frame(
  sample_id   = pop.ids,
  bio_label   = NEH.labels,
  stringsAsFactors = FALSE
)

# Derive treatment from the biological label
meta$treatment <- ifelse(grepl("_CB", meta$bio_label), "control",
                  ifelse(grepl("_SB", meta$bio_label), "stress", "initial"))
# can also parse by trial  
meta$trial <- sub("_.*", "", meta$bio_label)
# and parse by population
meta$pop <- ifelse(grepl("^ARC", meta$bio_label), "ARC",
                   ifelse(grepl("^NEH",  meta$bio_label),  "NEH",
                          ifelse(grepl("^MV", meta$bio_label), "MV", NA)))
View(meta)

#check that column names in input.all.rand.af and sample_id column in meta table are the same
all.equal(colnames(NEH.input.all.neutral.af), meta$sample_id)
# should be TRUE

# assign the columns of input.all.neutral.af with their proper sample names instead of the sample_id created by inputting the table to poolseq
colnames(NEH.input.all.neutral.af) <- meta$bio_label
```

```{r}
# Perform PCA
af_t <- t(NEH.input.all.neutral.af)

# Run PCA on transposed allele frequency matrix (samples as rows, loci as columns)
pca_res <- prcomp(af_t, center = TRUE, scale. = TRUE)

# Get scores for plotting
pca_df <- as.data.frame(pca_res$x)
#pca_df$locus <- rownames(input.all.neutral.af)  # keep locus IDs if you have them

# Calculate percent variance explained by each PC
var_explained <- pca_res$sdev^2 / sum(pca_res$sdev^2) * 100

pca_df$Treatment <- meta$treatment
pca_df$Trial <- meta$trial

neutpca_neh <- ggplot(pca_df, aes(x = PC1, y = PC2, color = Treatment, fill=Treatment)) +
  geom_point(aes(alpha=0.2), size =7,shape=21, col="black")+  
  scale_fill_manual(values=treatPalette,name="Treatment")+ 
  annotate("text", x = -330, y = -40, label = "NEH1", color = "black", size = 5) +
  annotate("text", x = 120, y = -250, label = "NEH2", color = "black", size = 5) +
  annotate("text", x = 150, y = 170, label = "NEH3", color = "black", size = 5) +
guides(alpha=FALSE) +theme(axis.title.x = element_text(size = 24),axis.title.y = element_text(size = 24)) +
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 1)
    ) +
  labs(
    x = paste0("PC1 (", round(var_explained[1], 1), "%)"),
    y = paste0("PC2 (", round(var_explained[2], 1), "%)"),
    title = "PCA of Neutral Loci (NEH)"
  )
neutpca_neh
ggsave(filename = "PCA/neutral_PCA_NEH.png", plot = neutpca_neh, width = 6, height = 5, units = "in", dpi = 300)
```


### MV
```{bash, eval = FALSE}
# Activate the CASE conda environment
source activate CADO

mawk '!/CHR/' MV.neutral.only.sync | cut -f1,2 | sort | uniq > neutral.pca.filter.MV

mawk 'NR==FNR{a[$1,$2]; next} ($1,$2) in a' neutral.pca.filter.MV MV.dp25.cov.sync | cut --complement -f31- > MV.neutral.pca.sync
```

```{r}
# Read sync file of significant neutral loci for neutral PCA
MV.input.all.neutral <- read.sync(file="MV.neutral.pca.sync", gen=seq(1,27), repl=seq(1,27), polarization="rising")
MV.input.all.neutral.af <- af(MV.input.all.neutral,gen=seq(1,27), repl=seq(1,27))

MV.labels <- c(
  "MV1_CB1","MV1_CB2","MV1_CB3","MV1_SB1","MV1_SB2","MV1_SB3","MV1_T01","MV1_T02","MV1_T03",
  "MV2_CB1","MV2_CB2","MV2_CB3","MV2_SB1","MV2_SB2","MV2_SB3","MV2_T01","MV2_T02","MV2_T03",
  "MV3_CB1","MV3_CB2","MV3_CB3","MV3_SB1","MV3_SB2","MV3_SB3","MV3_T01","MV3_T02","MV3_T03"
)
length(MV.labels)  # should be 27

pop.ids <- colnames(MV.input.all.neutral.af)  # "F1.R1.freq", ..., "F63.R63.freq"

meta <- data.frame(
  sample_id   = pop.ids,
  bio_label   = MV.labels,
  stringsAsFactors = FALSE
)

# Derive treatment from the biological label
meta$treatment <- ifelse(grepl("_CB", meta$bio_label), "control",
                  ifelse(grepl("_SB", meta$bio_label), "stress", "initial"))
# can also parse by trial  
meta$trial <- sub("_.*", "", meta$bio_label)
# and parse by population
meta$pop <- ifelse(grepl("^ARC", meta$bio_label), "ARC",
                   ifelse(grepl("^NEH",  meta$bio_label),  "NEH",
                          ifelse(grepl("^MV", meta$bio_label), "MV", NA)))
View(meta)

#check that column names in input.all.rand.af and sample_id column in meta table are the same
all.equal(colnames(MV.input.all.neutral.af), meta$sample_id)
# should be TRUE

# assign the columns of input.all.neutral.af with their proper sample names instead of the sample_id created by inputting the table to poolseq
colnames(MV.input.all.neutral.af) <- meta$bio_label
```

```{r}
# Perform PCA
af_t <- t(MV.input.all.neutral.af)

# Run PCA on transposed allele frequency matrix (samples as rows, loci as columns)
pca_res <- prcomp(af_t, center = TRUE, scale. = TRUE)

# Get scores for plotting
pca_df <- as.data.frame(pca_res$x)
#pca_df$locus <- rownames(input.all.neutral.af)  # keep locus IDs if you have them

# Calculate percent variance explained by each PC
var_explained <- pca_res$sdev^2 / sum(pca_res$sdev^2) * 100

pca_df$Treatment <- meta$treatment
pca_df$Trial <- meta$trial

neutpca_mv <- ggplot(pca_df, aes(x = PC1, y = PC2, color = Treatment, fill=Treatment)) +
  geom_point(aes(alpha=0.2), size =7,shape=21, col="black")+  
  scale_fill_manual(values=treatPalette,name="Treatment")+ 
  annotate("text", x = -180, y = 280, label = "MV1", color = "black", size = 5) +
  annotate("text", x = -200, y = -250, label = "MV2", color = "black", size = 5) +
  annotate("text", x = 500, y = 50, label = "MV3", color = "black", size = 5) +
guides(alpha=FALSE) +theme(axis.title.x = element_text(size = 24),axis.title.y = element_text(size = 24)) +
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 1)
    ) +
  labs(
    x = paste0("PC1 (", round(var_explained[1], 1), "%)"),
    y = paste0("PC2 (", round(var_explained[2], 1), "%)"),
    title = "PCA of Neutral Loci (MV)"
  )
neutpca_mv
ggsave(filename = "PCA/neutral_PCA_MV.png", plot = neutpca_mv, width = 6, height = 5, units = "in", dpi = 300)
```

### Neutral PCA Composite Figure
```{r}
neutpca_design <- "AAABBBCCC"

allpop_neutpca <- (neutpca_arc + neutpca_neh + neutpca_mv) +
  plot_layout(design = neutpca_design, guides = "collect") +
  plot_annotation(tag_levels = "a", tag_suffix = ")") 
print(allpop_neutpca)

# Export to PNG at a specific size
ggsave("PCA/neutral_pca_allpopcomposite.png", plot = allpop_neutpca, width = 15, height =5, units = "in", dpi = 300)
```