---
title: "Ch 1 Larval CADO w/ multiple populations"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = TRUE)
```

### Setup
```{r include=TRUE}
library(ggplot2)
library(tidyr)
library(plyr)
library(dplyr)
library(qvalue)
library(stringr)
library(pcadapt)
library(poolSeq)
library(ACER)
library(ggman)
library(VennDiagram)
library(scales)
library(data.table)
library(patchwork)
```

```{r include=TRUE}
# Function to convert p-values to q-values
Qvalue_convert <- function(table) {


table <- spread(table,GROUP,PVAL)  
  
table$CHR <- table$CHROM

table %>% 
  mutate(CHR = str_replace(CHR, "NC_035780.1", "1")) %>% 
  mutate(CHR = str_replace(CHR, "NC_035781.1", "2")) %>% 
  mutate(CHR = str_replace(CHR, "NC_035782.1", "3")) %>% 
  mutate(CHR = str_replace(CHR, "NC_035783.1", "4")) %>% 
  mutate(CHR = str_replace(CHR, "NC_035784.1", "5")) %>% 
  mutate(CHR = str_replace(CHR, "NC_035785.1", "6")) %>% 
  mutate(CHR = str_replace(CHR, "NC_035786.1", "7")) %>% 
  mutate(CHR = str_replace(CHR, "NC_035787.1", "8")) %>%
  mutate(CHR = str_replace(CHR, "NC_035788.1", "9")) %>% 
  mutate(CHR = str_replace(CHR, "NC_035789.1", "10"))  -> pp1

pp1 <- unite(pp1, "SNP", c("CHROM","BP"),remove = FALSE)

pp1$CHR <- as.numeric(pp1$CHR)  


# Set pi0=1 to equal Benjamini Hochberg
pp1$QCC <- qvalue(pp1$PCC, pi0 = 1)$qvalues
pp1$QCS <- qvalue(pp1$PCS, pi0 = 1)$qvalues
pp1$QVC <- qvalue(pp1$PVC, pi0 = 1)$qvalues
pp1$QVS <- qvalue(pp1$PVS, pi0 = 1)$qvalues


return(pp1)
}
```

```{r include=TRUE}
# Function to determine significant loci in various outputs
Significant_subset <- function(pv, alpha, alpha2) {
  # Add new logical columns based on conditions
  #pv$Sig.CC <- pv$QCC < alpha #not including this one treatment as it's the true "control"
  pv$Sig.CS <- pv$QCS < alpha
  pv$Sig.VC <- pv$QVC < alpha
  pv$Sig.VS <- pv$QVS < alpha
  
  # Subset the dataframe
  ppsig <- subset(pv, QCS < alpha | QVC < alpha | QVS < alpha )
  ppsig <- subset(ppsig, QCC > alpha2)
  
  # Print diagnostics
  print(nrow(pv))
  print(nrow(ppsig))
  print(nrow(ppsig) / nrow(pv))
  
  # Return the modified dataframe
  return(ppsig)
}
```

```{r include=TRUE}
group_and_average <- function(df) {
  df %>%
    dplyr::group_by(SNP) %>%
    dplyr::summarize(
      # Use a conditional inside across to handle all-NA groups
      across(
        c(PVC, PVS, PCC, PCS, CHR, QVC, QCC, QCS, QVS), 
        ~ if (all(is.na(.))) NA else min(., na.rm = TRUE)),
      # Combine logical columns, retain TRUE if any TRUE exists in the group
      Sig.VS = any(Sig.VS, na.rm = TRUE),
      Sig.VC = any(Sig.VC, na.rm = TRUE),
      Sig.CS = any(Sig.CS, na.rm = TRUE),
      CHROM = first(CHROM),
      BP = first(BP),
      .groups = "drop"  # Ungroup the result
    )
}
```

```{r include=TRUE}
#cbPaletteSmall <- c("purple", "green", "gold", "red", "orange", "blue", "violet")
cbPaletteSmall <- c("#3F8F7A", "#D3DDF0", "#4169E1", "#5F7BAA", "#F3D6D9", "#8C3A41", "hotpink")
cbPaletteSmall3 <- c("#3F8F7A", "#D8848A", "#5F7BAA")

# CC, CS, VC, VS
treatPalette <- c("#A29BAD", "white", "#D29F13", "#8CDFA0", "#EEF56E")
```

```{r}
sessionInfo()
```


# Visualize across 10,000 random loci
## All populations PCA of random 10k loci
```{bash, eval = FALSE}
mamba activate CADO

cut -f1,2 Ch2.dp25.cov.sync | sort -u  > filtered.loci

mawk '!/CHR/' filtered.loci | shuf -n 50000  | shuf -n 10000 > rand.loci

mawk 'NR==FNR{a[$1,$2]; next} ($1,$2) in a' rand.loci Ch2.dp25.cov.sync | cut --complement -f32- > input.all.rand.sync
```

```{r}
input.all.rand <- read.sync(file="input.all.rand.sync", gen=seq(1,28), repl=seq(1,28), polarization="rising")


input.all.rand.af <- af(input.all.rand,gen=seq(1,28), repl=seq(1,28))

bio.labels <- c(
  "CC1_TVECombined", "CC1_TVEVibrio",  "CC2_TVECombined","CC2_TVEVibrio",  "CC3_TVECombined", "CC3_TVEVibrio",  "CON_TV0Combined","CON_TV0Vibrio",  
  "CS1_TVECombined", "CS1_TVEVibrio",  "CS2_TVECombined", "CS2_TVEVibrio",  "CS3_TVECombined", "CS3_TVEVibrio",  "STR_TV0Combined", "STR_TV0Vibrio",  
  "VC1_TVECombined", "VC1_TVEVibrio",  "VC2_TVECombined","VC2_TVEVibrio",  "VC3_TVECombined",  "VC3_TVEVibrio","VS1_TVECombined","VS1_TVEVibrio", 
  "VS2_TVECombined", "VS2_TVEVibrio", "VS3_TVECombined", "VS3_TVEVibrio"
)
length(bio.labels)  # should be 28

pop.ids <- colnames(input.all.rand.af)  # "F1.R1.freq", ..., "F63.R63.freq"

meta <- data.frame(
  sample_id   = pop.ids,
  bio_label   = bio.labels,
  stringsAsFactors = FALSE
)

# Derive treatment from the biological label
meta$treatment <- ifelse(grepl("CC", meta$bio_label), "control_only",
                  ifelse(grepl("CS", meta$bio_label), "stress_only",
                  ifelse(grepl("VC", meta$bio_label), "vibrio_only",
                  ifelse(grepl("VS", meta$bio_label), "vibrio+stress", "initial"))))

#can also parse by probe set
meta$probes <- ifelse(grepl("Combined", meta$bio_label), "CADO+Vibrio",
                  ifelse(grepl("Vibrio", meta$bio_label), "Vibrio", ""))

View(meta)

#check that column names in input.all.rand.af and sample_id column in meta table are the same
all.equal(colnames(input.all.rand.af), meta$sample_id)
# should be TRUE

# assign the columns of input.all.rand.af with their proper sample names instead of the sample_id created by inputting the table to poolseq
colnames(input.all.rand.af) <- meta$bio_label
```

No real pattern to be seen! 
```{r}
# Perform PCA
af_t <- t(input.all.rand.af)

pca_res <- prcomp(af_t, center = TRUE, scale. = TRUE)

# Get scores for plotting
pca_df <- as.data.frame(pca_res$x)
#pca_df$locus <- rownames(input.all.rand.af)  # keep locus IDs if you have them

var_explained <- pca_res$sdev^2 / sum(pca_res$sdev^2) * 100

pca_df$Treatment <- meta$treatment
pca_df$Probes <- meta$probes

pca_rand10k <- ggplot(pca_df, aes(x = PC1, y = PC2, color = Treatment, fill=Treatment)) +
  geom_point(aes(alpha=1), size =4,shape=21, col="black")+  scale_fill_manual(values=treatPalette,name="Treatment")+ 
guides(alpha=FALSE) +theme(axis.title.x = element_text(size = 24),axis.title.y = element_text(size = 24)) +
  theme_minimal() +
  labs(
    x = paste0("PC1 (", round(var_explained[1], 1), "%)"),
    y = paste0("PC2 (", round(var_explained[2], 1), "%)"),
    title = "PCA of Allele Frequencies - random 10K loci"
  ); pca_rand10k

ggsave(filename = "random_PCA/ch1_pca_rand10k.png", plot = pca_rand10k, width = 6, height = 5, units = "in", dpi = 300)
```

### PCA of random 10K loci from All Initial Samples
```{r}
#subset data to only T0 initial samples
t0_cols <- grepl("TV0", colnames(input.all.rand.af))
input.T0.rand.af <- input.all.rand.af[, t0_cols]
afi_t <- t(input.T0.rand.af)

#subset metadata table
t0_meta <- subset(meta, meta$treatment == "initial")
View(t0_meta)

# start pca code
zero_var_cols <- apply(afi_t, 2, var, na.rm = TRUE) == 0
afi_t_filtered <- afi_t[, !zero_var_cols]
pca_i_res <- prcomp(afi_t_filtered, center = TRUE, scale. = TRUE)

# Get scores for plotting
pca_i_df <- as.data.frame(pca_i_res$x)

var_explained <- pca_i_res$sdev^2 / sum(pca_i_res$sdev^2) * 100

pca_i_df$Treatment <- t0_meta$treatment
pca_i_df$Probes <- t0_meta$probes
pca_i_df$Label <- t0_meta$bio_label

pca_t0rand10k<- ggplot(pca_i_df, aes(x = PC1, y = PC2, color = Label, fill=Label)) +
  geom_point(aes(alpha=0.2), size =6,shape=21, col="black")+  scale_fill_manual(values=cbPaletteSmall,name="Treatment")+ 
guides(alpha=FALSE) +theme(axis.title.x = element_text(size = 24),axis.title.y = element_text(size = 24)) +
  theme_classic() +
  labs(
    x = paste0("PC1 (", round(var_explained[1], 1), "%)"),
    y = paste0("PC2 (", round(var_explained[2], 1), "%)"),
    title = "PCA of Allele Frequencies"
  );pca_t0rand10k

ggsave(filename = "random_PCA/ch1_pca_initial_rand10k.png", plot = pca_t0rand10k, width = 6, height = 5, units = "in", dpi = 300)
```


#### Random loci for each individual experimental population (ARC, NEH, MV)
```{r}
# Ensure meta is keyed by bio_label and in column order
meta <- meta[match(colnames(input.all.rand.af), meta$bio_label), ]
stopifnot(all(meta$bio_label == colnames(input.all.rand.af)))

pops <- c("ARC", "MV", "NEH")

for (pop in pops) {
  # select columns belonging to this population
  cols_idx <- which(meta$pop == pop)
  cat(sprintf("\n== Population %s: found %d sample columns ==\n", pop, length(cols_idx)))
  if (length(cols_idx) < 2) {
    warning(sprintf("Population %s: less than 2 samples found - skipping PCA.", pop))
    next
  }

  sub_mat <- input.all.rand.af[, cols_idx, drop = FALSE]
  cat("  dim(sub_mat) (loci x samples):", dim(sub_mat), "\n")

  # transpose: rows = samples, cols = loci
  mat_t <- t(sub_mat)

  # drop zero-variance loci within this population
  var_cols <- apply(mat_t, 2, var, na.rm = TRUE)
  keep     <- which(var_cols > 0)
  removed_n <- sum(var_cols == 0)
  cat("  loci removed (zero variance):", removed_n, "\n")

  if (length(keep) < 2) {
    warning(sprintf("Population %s: fewer than 2 variable loci after filtering - skipping PCA.", pop))
    next
  }
  mat_t2 <- mat_t[, keep, drop = FALSE]
  cat("  dim(mat_t2) (samples x retained loci):", dim(mat_t2), "\n")

  # PCA
  pca_res <- prcomp(mat_t2, center = TRUE, scale. = TRUE)
  pca_df  <- as.data.frame(pca_res$x)

  # add metadata for these samples
  sample_labels <- rownames(mat_t2)                      # bio_labels
  sample_meta   <- meta[match(sample_labels, meta$bio_label), ]

  pca_df$population <- sample_meta$population            # should all be 'pop'
  pca_df$treatment  <- sample_meta$treatment             # T0 / CB / SB
  pca_df$bio_label  <- sample_meta$bio_label
  pca_df$trial <- sample_meta$trial

  var_explained <- (pca_res$sdev^2) / sum(pca_res$sdev^2) * 100

  # palette for treatment or population; here color by treatment
  if (exists("treatPalette")) {
    fill_scale <- scale_fill_manual(values = treatPalette, name = "Treatment")
  } else {
    fill_scale <- scale_fill_brewer(palette = "Set1", name = "Treatment")
  }

  # plot: within a population, color by treatment
  p <- ggplot(pca_df, aes(x = PC1, y = PC2, fill = treatment)) +
    geom_point(size = 5, shape = 21, color = "black", alpha = 0.8) +
    fill_scale +
    geom_text_repel(aes(label = trial), size = 3) +
    theme_classic() +
    theme(axis.title.x = element_text(size = 14),
          axis.title.y = element_text(size = 14),
          plot.title   = element_text(hjust = 0.5)) +
    labs(
      title = paste0("PCA - Population ", pop),
      x = paste0("PC1 (", round(var_explained[1], 1), "%)"),
      y = paste0("PC2 (", round(var_explained[2], 1), "%)")
    )

  outfile <- paste0("PCA/PCA_population_", pop, ".png")
  ggsave(outfile, plot = p, width = 10, height = 7, dpi = 300)
  cat("  saved:", outfile, "\n")
}

```

#### Random loci for each individual experimental trial (ARC2, NEH1-3, MV1-3)
```{r}
# Ensure meta is keyed by bio_label and in column order
meta <- meta[match(colnames(input.all.rand.af), meta$bio_label), ]
stopifnot(all(meta$bio_label == colnames(input.all.rand.af)))

trials <- c("ARC2", "MV1", "MV2", "MV3", "NEH1", "NEH2", "NEH3")

for (trial in trials) {
  # select columns belonging to this population
  cols_idx <- which(meta$trial == trial)
  cat(sprintf("\n== Population %s: found %d sample columns ==\n", trial, length(cols_idx)))
  if (length(cols_idx) < 2) {
    warning(sprintf("Population %s: less than 2 samples found - skipping PCA.", trial))
    next
  }

  sub_mat <- input.all.rand.af[, cols_idx, drop = FALSE]
  cat("  dim(sub_mat) (loci x samples):", dim(sub_mat), "\n")

  # transpose: rows = samples, cols = loci
  mat_t <- t(sub_mat)

  # drop zero-variance loci within this population
  var_cols <- apply(mat_t, 2, var, na.rm = TRUE)
  keep     <- which(var_cols > 0)
  removed_n <- sum(var_cols == 0)
  cat("  loci removed (zero variance):", removed_n, "\n")

  if (length(keep) < 2) {
    warning(sprintf("Trial %s: fewer than 2 variable loci after filtering - skipping PCA.", trial))
    next
  }
  mat_t2 <- mat_t[, keep, drop = FALSE]
  cat("  dim(mat_t2) (samples x retained loci):", dim(mat_t2), "\n")

  # PCA
  pca_res <- prcomp(mat_t2, center = TRUE, scale. = TRUE)
  pca_df  <- as.data.frame(pca_res$x)

  # add metadata for these samples
  sample_labels <- rownames(mat_t2)                      # bio_labels
  sample_meta   <- meta[match(sample_labels, meta$bio_label), ]

  pca_df$population <- sample_meta$population            # should all be 'pop'
  pca_df$treatment  <- sample_meta$treatment             # T0 / CB / SB
  pca_df$bio_label  <- sample_meta$bio_label
  pca_df$trial <- sample_meta$trial

  var_explained <- (pca_res$sdev^2) / sum(pca_res$sdev^2) * 100

  # palette for treatment or population; here color by treatment
  if (exists("treatPalette")) {
    fill_scale <- scale_fill_manual(values = treatPalette, name = "Treatment")
  } else {
    fill_scale <- scale_fill_brewer(palette = "Set1", name = "Treatment")
  }

  # plot: within a population, color by treatment
  p <- ggplot(pca_df, aes(x = PC1, y = PC2, fill = treatment)) +
    geom_point(size = 5, shape = 21, color = "black", alpha = 0.8) +
    fill_scale +
    theme_classic() +
    theme(axis.title.x = element_text(size = 14),
          axis.title.y = element_text(size = 14),
          plot.title   = element_text(hjust = 0.5)) +
    labs(
      title = paste0("PCA - Exp. Trial ", trial),
      x = paste0("PC1 (", round(var_explained[1], 1), "%)"),
      y = paste0("PC2 (", round(var_explained[2], 1), "%)")
    )

  outfile <- paste0("PCA/PCA_trial_", trial, ".png")
  ggsave(outfile, plot = p, width = 10, height = 7, dpi = 300)
  cat("  saved:", outfile, "\n")
}
```

#### Venn Diagram for called, filtered loci between populations
Haven't updated this code yet just adding in if I want to come back to it later.
Right now it doesn't seem like this will be that informative as most of my variation is driven by trial (7 groups)
```{bash}
# extract the CHROM and POS of snps from each sync file 
cut -f1,2 ARC.dp25.cov.sync > ARC.pos
cut -f1,2 NEH.dp25.cov.sync > NEH.pos
cut -f1,2 MV.dp25.cov.sync > MV.pos

# count overlap between trials 
cat <(echo "SNP") <(cat *.pos | cut -f1,2 | sort | uniq -c | mawk '$1 <2' | mawk '{print $2"_"$3}') > singleton.loci
```

```{r}
# Read and preprocess data
ARC.loci <- read.table("ARC.pos", header = TRUE)
colnames(ARC.loci) <- c("CHROM", "BP")
ARC.loci <- ARC.loci %>% mutate(ARC = TRUE)

NEH.loci <- read.table("NEH.pos", header = TRUE)
colnames(NEH.loci) <- c("CHROM", "BP")
NEH.loci <- NEH.loci %>% mutate(NEH = TRUE)

MV.loci <- read.table("MV.pos", header = TRUE)
colnames(MV.loci) <- c("CHROM", "BP")
MV.loci <- MV.loci %>% mutate(MV = TRUE)

# Full join all data frames to preserve all CHROM and BP combinations
pos_df <- ARC.loci %>%
  full_join(NEH.loci %>% select(CHROM, BP, NEH), by = c("CHROM", "BP")) %>%
  full_join(MV.loci %>% select(CHROM, BP, MV), by = c("CHROM", "BP"))

# Replace NA values in the BLOCK_* columns with FALSE
pos_df <- pos_df %>%
  mutate(
    ARC = ifelse(is.na(ARC), FALSE, ARC),
    NEH = ifelse(is.na(NEH), FALSE, NEH),
    MV = ifelse(is.na(MV), FALSE, MV)
  )


loci.arc <- nrow(subset(pos_df,ARC == TRUE))
loci.neh <- nrow(subset(pos_df,NEH == TRUE))
loci.mv <- nrow(subset(pos_df,MV == TRUE))

loci.arc.neh <- nrow(subset(pos_df,ARC == TRUE & NEH ==TRUE))
loci.arc.mv <- nrow(subset(pos_df,ARC == TRUE & MV ==TRUE))
loci.neh.mv <- nrow(subset(pos_df,NEH == TRUE & MV ==TRUE))
loci.arc.neh.mv <- nrow(subset(pos_df,ARC == TRUE & NEH ==TRUE & MV ==TRUE))

png(filename="Venn_variants.png", type="cairo",units="px", width=4000, height=4000, res=500, bg="transparent")

venn.plot <- draw.triple.venn(
  area1=loci.arc, area2=loci.neh, area3=loci.mv, 
  n12= loci.arc.neh, n13=loci.arc.mv, 
  n23=loci.neh.mv,scaled=TRUE,
  n123=loci.arc.neh.mv, cex=rep(2, 7),
  category = c("ARC", "NEH", "MV"),
  fill = cbPaletteSmall3, cat.cex = rep(2, 3),
  cat.col = cbPaletteSmall3, label.col = rep("white", 7))
dev.off()

```





# Population level outlier testing

## Vibrio Probed only
Vibrio probed samples were captured with probes generated from the mRNA of larval pools collected during this Vibrio challenge.

First normalize with initial sample coverage. The core logic is that the internal standard (IS) pools (the inital samples) act as a reference to normalize and cap coverage, ensuring that outlier detection downstream isn't driven by coverage artifacts rather than true allele frequency signals.

> NOTE: setting separate internal stanards - one for CON & one for STR!
> then later, I'll use the pre-treatment specific IS for each treatment group in the CMH

### Set CON internal standard with CON_TV0 sample
```{bash, eval=FALSE}
source activate CADO

# find positions of CON initial sample 
mawk 'NR==1 { for(i=1; i<=NF; i++) print i, $i }' VibrioProbed.dp25.cov.sync 
# CON is 7

#sum up initial samples - don't need to sum bc its only one sample, but this makes the right file!
bash ../Scripts/sum.sh <(cut -f1,2,3,7 VibrioProbed.dp25.cov.sync) > VibrioProbed.CON.IS.sync

# makes extra columns with the "sum" data
paste VibrioProbed.CON.IS.sync <(cut -f4 VibrioProbed.CON.IS.sync) <(cut -f4 VibrioProbed.CON.IS.sync)  > VibrioProbed.CON.IS.sum3.sync

# look at what col number to put in the compute average step next 
mawk 'NR==1 { for(i=1; i<=NF; i++) print i, $i }' VibrioProbed.dp25.cov.sync
# remove both CON AND STR TV0 samples and compute coverage stats
mawk -f ../Scripts/add_cov_sync.txt <(cut -f7,11 --complement VibrioProbed.dp25.cov.sync) > VibrioProbed.sansIS.cov.sync

mawk 'NR==1 { for(i=1; i<=NF; i++) print i, $i }' VibrioProbed.sansIS.cov.sync # last column is 21 (mean coverage w/o initial samples)

# compute average coverage on sync file minus initial samples
AV_COV=$(mawk -f ../Scripts/add_cov_sync.txt <(cut -f7,11 --complement VibrioProbed.dp25.cov.sync) | mawk '{sum=sum+$21} END {print sum/NR}')
echo $AV_COV

mawk -f ../Scripts/add_cov_sync.txt VibrioProbed.CON.IS.sum3.sync | mawk -v x=$AV_COV '$7 >= x' | mawk '!/CHR/' | cut -f1-6 > VibrioProbed.CON.IS.sum3nh.sync

mawk -f ../Scripts/add_cov_sync_IS.txt VibrioProbed.CON.IS.sum3.sync | mawk -v x=$AV_COV '$7 >= x'> VibrioProbed.CON.IS.sum3nh.cov.sync

# apply coverage filter back to main sync

#first check col number to apply filter to
paste VibrioProbed.dp25.cov.sync <(mawk -f ../Scripts/add_cov_sync.txt VibrioProbed.CON.IS.sum3.sync ) > CON.ISnormVibrioProbed.cov.sync
mawk 'NR==1 { for(i=1; i<=NF; i++) print i, $i }' CON.ISnormVibrioProbed.cov.sync # col 29 (mean cov normalized by initial samples)

# apply initial sample coverage filter
paste VibrioProbed.dp25.cov.sync <(mawk -f ../Scripts/add_cov_sync.txt VibrioProbed.CON.IS.sum3.sync ) | mawk -v x=$AV_COV '$29 >= x' | cut -f1-20 > temp.sync

# VibrioProbed samples normalized by CON initial sample coverage stored in *COV.sync
mv temp.sync VibrioProbed.CON.IS.dp25.COV.sync

# cap coverage at IS maxiumum
#remove inital samples and remove old cov stats and recalculates the coverage with only treatment pools
mawk -f ../Scripts/add_cov_sync.txt <(cut -f7,11,18- --complement VibrioProbed.CON.IS.dp25.COV.sync) > temp.cov

# pastes pop coverage with minimum inital sample coverage (col 7) to last col (19) if temp.cov and compare treatemtn pool mean cov ($12) to IS min cov ($13)
paste temp.cov <(cut -f7 VibrioProbed.CON.IS.sum3nh.cov.sync) | mawk -v OFS='\t' '{if ($18 > $19) {$18=$19; print $0} else {print $0}}' | cut -f1-18> VibrioProbed.CON.IS.dp25.COV
```
The final output is VibrioProbed.CON.IS.dp25.COV.sync, a fully processed, CON_TV0 coverage-normalized sync ready for outlier detection.

### Set STR internal standard with STR_TV0 sample
```{bash, eval=FALSE}
source activate CADO

# find positions of STR initial sample 
mawk 'NR==1 { for(i=1; i<=NF; i++) print i, $i }' VibrioProbed.dp25.cov.sync 
# STR is 11

#sum up initial samples - don't need to sum bc its only one sample, but this makes the right file!
bash ../Scripts/sum.sh <(cut -f1,2,3,11 VibrioProbed.dp25.cov.sync) > VibrioProbed.STR.IS.sync

# makes extra columns with the "sum" data
paste VibrioProbed.STR.IS.sync <(cut -f4 VibrioProbed.STR.IS.sync) <(cut -f4 VibrioProbed.STR.IS.sync)  > VibrioProbed.STR.IS.sum3.sync

# look at what col number to put in the compute average step next 
#mawk 'NR==1 { for(i=1; i<=NF; i++) print i, $i }' VibrioProbed.dp25.cov.sync
# remove both CON AND STR TV0 samples and compute coverage stats
#mawk -f ../Scripts/add_cov_sync.txt <(cut -f7,11 --complement VibrioProbed.dp25.cov.sync) > VibrioProbed.sansIS.cov.sync

#mawk 'NR==1 { for(i=1; i<=NF; i++) print i, $i }' VibrioProbed.sansIS.cov.sync # last column is 21 (mean coverage w/o initial samples)

# compute average coverage on sync file minus initial samples
#AV_COV=$(mawk -f ../Scripts/add_cov_sync.txt <(cut -f7,11 --complement VibrioProbed.dp25.cov.sync) | mawk '{sum=sum+$21} END {print sum/NR}')
echo $AV_COV

mawk -f ../Scripts/add_cov_sync.txt VibrioProbed.STR.IS.sum3.sync | mawk -v x=$AV_COV '$7 >= x' | mawk '!/CHR/' | cut -f1-6 > VibrioProbed.STR.IS.sum3nh.sync

mawk -f ../Scripts/add_cov_sync_IS.txt VibrioProbed.STR.IS.sum3.sync | mawk -v x=$AV_COV '$7 >= x'> VibrioProbed.STR.IS.sum3nh.cov.sync

# apply coverage filter back to main sync

#first check col number to apply filter to
paste VibrioProbed.dp25.cov.sync <(mawk -f ../Scripts/add_cov_sync.txt VibrioProbed.STR.IS.sum3.sync ) > STR.ISnormVibrioProbed.cov.sync
mawk 'NR==1 { for(i=1; i<=NF; i++) print i, $i }' STR.ISnormVibrioProbed.cov.sync # col 29 (mean cov normalized by initial samples)

# apply initial sample coverage filter
paste VibrioProbed.dp25.cov.sync <(mawk -f ../Scripts/add_cov_sync.txt VibrioProbed.STR.IS.sum3.sync ) | mawk -v x=$AV_COV '$29 >= x' | cut -f1-20 > temp.sync

# VibrioProbed samples normalized by STR initial sample coverage stored in *COV.sync
mv temp.sync VibrioProbed.STR.IS.dp25.COV.sync

# cap coverage at IS maxiumum
#remove inital samples and remove old cov stats and recalculates the coverage with only treatment pools
mawk -f ../Scripts/add_cov_sync.txt <(cut -f7,11,18- --complement VibrioProbed.STR.IS.dp25.COV.sync) > temp.cov

# pastes pop coverage with minimum inital sample coverage (col 7) to last col (19) if temp.cov and compare treatemtn pool mean cov ($12) to IS min cov ($13)
paste temp.cov <(cut -f7 VibrioProbed.STR.IS.sum3nh.cov.sync) | mawk -v OFS='\t' '{if ($18 > $19) {$18=$19; print $0} else {print $0}}' | cut -f1-18> VibrioProbed.STR.IS.dp25.COV
```
The final output is VibrioProbed.STR.IS.dp25.COV.sync, a fully processed, STR_TV0 coverage-normalized sync ready for outlier detection.

> NOTE: In the following steps the treatment data will be interleaved with the internal standard (initital sample) data. 
> CMH test uses the internal standard columns to set the null expectation during outlier detection. 
> For this project, my initial samples are from both CON and STR pools (prior exposure) so I will run separate CMH tests for each pre-treatment (CADO) exposure

```{bash, eval=FALSE}
source activate random_draw

# subsampling the internal standard - CON
python ../Scripts/sub_sample.py VibrioProbed.CON.IS.sum3nh.cov.sync VibrioProbed.CON.IS.sum3nh.sync VibrioProbed.CON.ISs.sync
# add heading
cat <(echo -e "CHROM\tPOS\tREF\tIS_RS1\tIS_RS2\tIS_RS3") VibrioProbed.CON.ISs.sync > VibrioProbed.CON.ISsum.sync 
#use this file below for con pre-treament samples (CC, VC)

# subsampling the internal standard - STR
python ../Scripts/sub_sample.py VibrioProbed.STR.IS.sum3nh.cov.sync VibrioProbed.STR.IS.sum3nh.sync VibrioProbed.STR.ISs.sync
# add heading
cat <(echo -e "CHROM\tPOS\tREF\tIS_RS1\tIS_RS2\tIS_RS3") VibrioProbed.STR.ISs.sync > VibrioProbed.STR.ISsum.sync  
#use this file below for stress pre-treatment samples (CS, VS)
```

Build treatment specific input files interleaved with the CON/STR internal standard (IS) columns;  IS columns will serve as the null expectation during outlier detection
```{bash, eval=FALSE}
# CC columns - will need to update columns based on data
paste <(cut -f1,2,3,4 VibrioProbed.CON.ISsum.sync) <(cut -f4 VibrioProbed.CON.IS.dp25.COV.sync) \
      <(cut -f5 VibrioProbed.CON.ISsum.sync) <(cut -f5 VibrioProbed.CON.IS.dp25.COV.sync) \
      <(cut -f6 VibrioProbed.CON.ISsum.sync) <(cut -f6 VibrioProbed.CON.IS.dp25.COV.sync)> VibrioProbed.CC.sync

head -1 VibrioProbed.CC.sync
tail -n +2 VibrioProbed.CC.sync > VibrioProbed.CC.input

# VC  - will need to update columns based on data
paste <(cut -f1,2,3,4 VibrioProbed.CON.ISsum.sync) <(cut -f12 VibrioProbed.CON.IS.dp25.COV.sync) \
      <(cut -f5 VibrioProbed.CON.ISsum.sync) <(cut -f13 VibrioProbed.CON.IS.dp25.COV.sync) \
      <(cut -f6 VibrioProbed.CON.ISsum.sync) <(cut -f14 VibrioProbed.CON.IS.dp25.COV.sync)> VibrioProbed.VC.sync

head -1 VibrioProbed.VC.sync
tail -n +2 VibrioProbed.VC.sync > VibrioProbed.VC.input

# CS columns - will need to update columns based on data
paste <(cut -f1,2,3,4 VibrioProbed.STR.ISsum.sync)  <(cut -f8 VibrioProbed.STR.IS.dp25.COV.sync) \
      <(cut -f5 VibrioProbed.STR.ISsum.sync) <(cut -f9 VibrioProbed.STR.IS.dp25.COV.sync) \
      <(cut -f6 VibrioProbed.STR.ISsum.sync) <(cut -f10 VibrioProbed.STR.IS.dp25.COV.sync)> VibrioProbed.CS.sync
      
head -1 VibrioProbed.CS.sync
tail -n +2 VibrioProbed.CS.sync > VibrioProbed.CS.input 

# VS columns - will need to update columns based on data
paste <(cut -f1,2,3,4 VibrioProbed.STR.ISsum.sync)  <(cut -f15 VibrioProbed.STR.IS.dp25.COV.sync) \
      <(cut -f5 VibrioProbed.STR.ISsum.sync) <(cut -f16 VibrioProbed.STR.IS.dp25.COV.sync) \
      <(cut -f6 VibrioProbed.STR.ISsum.sync) <(cut -f17 VibrioProbed.STR.IS.dp25.COV.sync)> VibrioProbed.VS.sync
      
head -1 VibrioProbed.VS.sync
tail -n +2 VibrioProbed.VS.sync > VibrioProbed.VS.input 
```

input files: 
- VibrioProbed.CC.input
- VibrioProbed.VC.input
- VibrioProbed.CS.input 
- VibrioProbed.VS.input

### Calculate P-values - Vibrio probe
#### Modified CMH - Vibrio Probe libraries

Reads in the interleaved IS+treatment sync files into R and runs an adapted Cochran-Mantel-Haenszel (CMH) test to identify loci where allele frequencies shift significantly between initial samples and treatment pools.

Reading and parsing files - read.sync() requires a headerless input file
```{r}
# gen assigns each column a "generation," where 0 = IS (baseline) and 1 = treatment pool
# repl  pairs each IS replicate with its corresponding treatment pool as replicate 1, 2, and 3
# coverage() and af() extract per-locus read depth and allele frequencies from the parsed sync object, which are passed directly to the test.

#CC
CC.vibrio.sync <- read.sync(file="VibrioProbed.CC.input", gen=c(0, 1, 0, 1, 0, 1), repl=c(1, 1, 2, 2, 3,3))
CC.vibrio.cov <- coverage(CC.vibrio.sync,gen=c(0, 1, 0, 1, 0, 1), repl=c(1, 1, 2, 2, 3,3))
CC.vibrio.af <- af(CC.vibrio.sync,gen=c(0, 1, 0, 1, 0, 1), repl=c(1, 1, 2, 2, 3,3))

#CS
CS.vibrio.sync <- read.sync(file="VibrioProbed.CS.input", gen=c(0, 1, 0, 1, 0, 1), repl=c(1, 1, 2, 2, 3,3))
CS.vibrio.cov <- coverage(CS.vibrio.sync,gen=c(0, 1, 0, 1, 0, 1), repl=c(1, 1, 2, 2, 3,3))
CS.vibrio.af <- af(CS.vibrio.sync,gen=c(0, 1, 0, 1, 0, 1), repl=c(1, 1, 2, 2, 3,3))

#VC
VC.vibrio.sync <- read.sync(file="VibrioProbed.VC.input", gen=c(0, 1, 0, 1, 0, 1), repl=c(1, 1, 2, 2, 3,3))
VC.vibrio.cov <- coverage(VC.vibrio.sync,gen=c(0, 1, 0, 1, 0, 1), repl=c(1, 1, 2, 2, 3,3))
VC.vibrio.af <- af(VC.vibrio.sync,gen=c(0, 1, 0, 1, 0, 1), repl=c(1, 1, 2, 2, 3,3))

#VS
VS.vibrio.sync <- read.sync(file="VibrioProbed.VS.input", gen=c(0, 1, 0, 1, 0, 1), repl=c(1, 1, 2, 2, 3,3))
VS.vibrio.cov <- coverage(VS.vibrio.sync,gen=c(0, 1, 0, 1, 0, 1), repl=c(1, 1, 2, 2, 3,3))
VS.vibrio.af <- af(VS.vibrio.sync,gen=c(0, 1, 0, 1, 0, 1), repl=c(1, 1, 2, 2, 3,3))
```

#### Running adapted CMH test
The adapted CMH is a replicated test of allele frequency change — it asks whether allele frequencies shifted consistently across all 3 IS→treatment replicate pairs. **The parameters matter a lot here:**
- Ne=rep(250, 3) — effective population size of 250 per replicate, used to correct for genetic drift
  - estimated Ne to be around 400 because this is 1000x less than starting pool size of 400,000 larv per bucket and 100x less than 40,000 larv per initial sample
- poolSize=rep(c(40000, 10000), ...) — pool sizes alternating between IS (40,000 larvae) and treatment (40,000 larvae), used for appropriate variance correction
- MeanStart=TRUE — uses the mean IS allele frequency across replicates as the baseline rather than each IS independently
- mincov=20 — excludes loci with coverage below 20x in any pool, preventing low-coverage noise from driving false positives
```{r}
#CC
CC.vibrio.pvals <- adapted.cmh.test(freq=CC.vibrio.af, coverage=CC.vibrio.cov,
                                  Ne=rep(250, 3), gen=c(0, 1, 0, 1, 0, 1), repl=c(1, 1, 2, 2, 3,3),
                                  poolSize=rep(c(20000,20000), ncol(CC.vibrio.af)/2), MeanStart = TRUE, mincov =10)

p.CC.vibdf <- as.data.table(cbind(row.names(CC.vibrio.af),CC.vibrio.pvals))
colnames(p.CC.vibdf) <- c("Locus","PVAL")
p.CC.vibdf$GROUP <- "PCC"

#CS
CS.vibrio.pvals <- adapted.cmh.test(freq=CS.vibrio.af, coverage=CS.vibrio.cov,
                                  Ne=rep(250, 3), gen=c(0, 1, 0, 1, 0, 1), repl=c(1, 1, 2, 2, 3,3),
                                  poolSize=rep(c(20000,20000), ncol(CS.vibrio.af)/2), MeanStart = TRUE, mincov =10)

p.CS.vibdf <- as.data.table(cbind(row.names(CS.vibrio.af),CS.vibrio.pvals))
colnames(p.CS.vibdf) <- c("Locus","PVAL")
p.CS.vibdf$GROUP <- "PCS"

#VC
VC.vibrio.pvals <- adapted.cmh.test(freq=VC.vibrio.af, coverage=VC.vibrio.cov,
                                  Ne=rep(250, 3), gen=c(0, 1, 0, 1, 0, 1), repl=c(1, 1, 2, 2, 3,3),
                                  poolSize=rep(c(20000,20000), ncol(VC.vibrio.af)/2), MeanStart = TRUE, mincov =10)

p.VC.vibdf <- as.data.table(cbind(row.names(VC.vibrio.af),VC.vibrio.pvals))
colnames(p.VC.vibdf) <- c("Locus","PVAL")
p.VC.vibdf$GROUP <- "PVC"

#VS
VS.vibrio.pvals <- adapted.cmh.test(freq=VS.vibrio.af, coverage=VS.vibrio.cov,
                                  Ne=rep(250, 3), gen=c(0, 1, 0, 1, 0, 1), repl=c(1, 1, 2, 2, 3,3),
                                  poolSize=rep(c(20000,20000), ncol(VS.vibrio.af)/2), MeanStart = TRUE, mincov =10)

p.VS.vibdf <- as.data.table(cbind(row.names(VS.vibrio.af),VS.vibrio.pvals))
colnames(p.VS.vibdf) <- c("Locus","PVAL")
p.VS.vibdf$GROUP <- "PVS"


# build p value
Vib.pval.table <- bind_rows(p.CC.vibdf, p.CS.vibdf, p.VC.vibdf, p.VS.vibdf)
Vib.pval.table <- Vib.pval.table %>%
  separate(Locus, into = c("CHROM", "BP"), sep = "\\.(?=[^.]+$)")
Vib.pval.table$PVAL <- as.numeric(Vib.pval.table$PVAL)
```


#### convert p values to q values
This applies FDR (false discovery rate) correction to your CMH p-values for each population, converting raw p-values into q-values
```{r}
Qvalue_convert(Vib.pval.table) -> Vib.pv
```


## Outlier filtering - 10% FDR, 5% FDR and 1% FDR datasets - we'll use the 1% dataset
identify high-confidence candidate loci under selection due to stress across three populations
```{r}
#pp1 <- rbind(ARC.pv,NEH.pv,MV.pv)

# Filter significant loci at FDR 10% threshold; also apply top 10% quantile filter within significant loci
alpha = 0.1
alpha2 =0.1

ppsig.Vib.1 <- Significant_subset(Vib.pv,alpha,alpha2)
ppsig.Vib.1$PROBE <- "Vibrio"


#ppsig3.FDR10 <- bind_rows(ppsig.ARC.1,ppsig.NEH.1,ppsig.MV.1)
#all_sig.STR<- subset(ppsig3.FDR10, Sig.STR == "TRUE") %>% group_by(SNP) %>% filter(n()>2)
#all_sig.STR$Sig.CON<- FALSE
#write.table(all_sig.STR, "Sig.Loci.3", sep="\t", row.names = FALSE, quote = FALSE)
```


```{r}
# Filter significant loci at FDR 5% threshold; also apply top 10% quantile filter within significant loci
alpha = 0.05
alpha2 =0.1

ppsig.Vib.05 <- Significant_subset(Vib.pv,alpha,alpha2)

ppsig.Vib.05$PROBE <- "Vibrio"


#ppsig3.FDR05 <- rbind(ppsig.ARC.05,ppsig.NEH.05,ppsig.MV.05)
#multi_sig.STR<- subset(ppsig3.FDR05, Sig.STR == "TRUE") %>% group_by(SNP) %>% filter(n()>1)
#multi_sig.STR$Sig.CON <- FALSE
#multi_sig.CA<- subset(ppsig3.FDR05, QCA < NEH.alpha) %>% group_by(SNP) %>% filter(n()>1)
#multi_sig.CASE<- subset(ppsig3.FDR05, QCASE < NEH.alpha) %>% group_by(SNP) %>% filter(n()>1)
#multi_sig.SE<- subset(ppsig3.FDR05, QSE < NEH.alpha) %>% group_by(SNP) %>% filter(n()>1)
#multi_sig <- rbind(ppsig3.FDR05 %>% group_by(SNP) %>% filter(n()>1))
#multi_sig <- rbind(multi_sig.CA,multi_sig.CASE,multi_sig.SE)
#write.table(multi_sig.STR, "Sig.Loci.2", sep="\t", row.names = FALSE, quote = FALSE)
```

```{r}
#Filter significant loci at 1% FDR threshold; and take the top 10% 
alpha = 0.01
alpha2 =0.1

ppsig.Vib.01 <- Significant_subset(Vib.pv,alpha,alpha2)

```

```{r}
#conpp1 <- subset(pp1, QCC < alpha) - hold over from combining treatments

ppsig.Vib.01$PROBE <- "Vibrio"

write.table(ppsig.Vib.01, "VibrioProbe.Sig.Loci.FDR01.1", sep="\t", row.names = FALSE, quote = FALSE)
```

** I have filtered down to FDR of 1% **

#### extra filtering steps that Jon
I don't think I need these - they were used when Jon was building his all blocks total loci file. 
For individual population analysis I don't think I need to include this addition filter. 
```{r}
# don't need this extra filtering - this would be used if I was combining the populations and analyzing them together
# multi_sig.STR.01<- subset(ppsig3.FDR01, Sig.STR == "TRUE") %>% group_by(SNP) %>% filter(n()>1)
# multi_sig.STR.01$Sig.CON <- FALSE
# 
# write.table(multi_sig.STR.01, "Sig.Loci.FDR01.2", sep="\t", row.names = FALSE, quote = FALSE)
# 
# alpha = 0.01
# alpha2 =0.1
# NEH.alpha = alpha * 2
# 
# 
# ppsig.ARC <- Significant_subset(ARC.pv,alpha,alpha2)
#ppsig.ARC <- subset(ppsig.ARC, Sig.STR == TRUE & QSTR < quantile(ppsig.ARC$QSTR, na.rm = TRUE, probs = 0.01)
#                     )
# ppsig.NEH <- Significant_subset(NEH.pv,NEH.alpha,alpha2)
# ppsig.NEH <- subset(ppsig.NEH, Sig.STR == TRUE & QSTR < quantile(ppsig.NEH$QSTR, na.rm = TRUE, probs = 0.01)
#                     )
# ppsig.MV <- Significant_subset(MV.pv,alpha,alpha2)
#ppsig.MV <- subset(ppsig.MV, Sig.STR == TRUE & QSTR < quantile(ppsig.MV$QSTR, na.rm = TRUE, probs = 0.01)
#                     )
# 
# ppsig.ARC$POP <- "ARC"
# ppsig.NEH$POP <- "NEH"
# ppsig.MV$POP <- "MV"
# 
# singleton <- read.table("singleton.loci",header=TRUE)
# 
# ppsig1.s <- semi_join(rbind(ppsig.ARC,ppsig.NEH,ppsig.MV),singleton, by = "SNP")
```


## Control filtering
```{r}
conpp.Vibrio <- subset(Vib.pv, QCC < 0.001)
allcon.Vibrio <- subset(conpp.Vibrio, QCC / QCS < 10 & QCC / QVC < 10 & QCC/ QVS < 10)
write.table(allcon.Vibrio, "Vibrio.ALL.CON.SNPS", sep="\t", row.names=FALSE, quote=FALSE)
```

```{r}
total.sig.Vibrio <- anti_join(ppsig.Vib.01, allcon.Vibrio, by = "SNP")

write.table(total.sig.Vibrio, "Total.Significant.uf.Loci.Vibrio", sep="\t", row.names=FALSE, quote=FALSE)
```   

```{bash}
source activate CADO

for POP in Vibrio; do

  # Build per-population significant loci BED
  cut -f2,3 Total.Significant.uf.Loci.${POP} | tail -n +2 | sort -k1,2 | uniq \
    | mawk '{print $1 "\t" $2-1 "\t" $2}' \
    | bedtools sort -i - > Vibrio.Total.Significant.uf.loci.${POP}.bed

  # Build per-population CON loci gene list
  bedtools intersect -wb \
    -a <(mawk '{print $2"\t"$3-1"\t"$3}' ${POP}.ALL.CON.SNPS) \
    -b /home/Genomic_Resources/C_virginica/bed_and_GFF/sorted.ref3.0.gene.bed \
    | grep -oh "gene=LOC.*;g" | sed 's/gene=//g' | sed 's/;g//g' \
    | sort | uniq > CON.LOCI.${POP}

  # Build per-population CON filter BED
  bedtools intersect -wb \
    -a <(mawk '{print $2"\t"$3-1"\t"$3}' ${POP}.ALL.CON.SNPS) \
    -b /home/Genomic_Resources/C_virginica/bed_and_GFF/sorted.ref3.0.gene.bed \
    | cut -f4- | bedtools sort | bedtools merge > con.loci.filter.${POP}.bed

  # Build per-population ToFilter file
  cat <(echo -e "SNP\tCHROM\tBP") \
    <(bedtools intersect \
        -a Vibrio.Total.Significant.uf.loci.${POP}.bed \
        -b con.loci.filter.${POP}.bed \
      | mawk '{print $1"_"$3"\t"$1"\t"$3}') \
    > ToFilter.Con.Loci.${POP}.txt

done
```


```{r}
library(dplyr)

pops <- "Vibrio"

for (pop in pops) {

  con_filter <- read.table(paste0("ToFilter.Con.Loci.", pop, ".txt"), header = TRUE)

  total.sig <- get(paste0("total.sig.", pop))

  # Remove contaminating loci
  total.sig <- unique(anti_join(total.sig, con_filter, by = "SNP"))

  write.table(total.sig,
              paste0("Total.Significant.Loci.", pop),
              sep = "\t", row.names = FALSE,  quote = FALSE)
}
```


#### Can do BED conversion here and GO enrichment, but skipping for now




# Outlier PCA 
PCAs on outliers from different populations

## Vibrio Probes
```{bash, eval = FALSE}
# Activate the CASE conda environment
source activate CADO

mawk '!/CHR/' Total.Significant.Loci.Vibrio | cut -f2,3 | sort | uniq > TSL.pca.filter.Vibrio

mawk 'NR==FNR{a[$1,$2]; next} ($1,$2) in a' TSL.pca.filter.Vibrio VibrioProbed.dp25.cov.sync | cut --complement -f18- > Vibrio.outlier.pca.sync
```

```{r}
# Read sync file of significant outlier loci for outlier PCA
Vibrio.input.all.outlier <- read.sync(file="Vibrio.outlier.pca.sync", gen=seq(1,14), repl=seq(1,14), polarization="rising")
Vibrio.input.all.outlier.af <- af(Vibrio.input.all.outlier,gen=seq(1,14), repl=seq(1,14))

Vibrio.labels <- c(
  "CC1_TVEVibrio",   "CC2_TVEVibrio",   "CC3_TVEVibrio",   "CON_TV0Vibrio", 
  "CS1_TVEVibrio",   "CS2_TVEVibrio",  "CS3_TVEVibrio",   "STR_TV0Vibrio",
  "VC1_TVEVibrio",   "VC2_TVEVibrio",   "VC3_TVEVibrio",   "VS1_TVEVibrio",
  "VS2_TVEVibrio",   "VS3_TVEVibrio"
)
length(Vibrio.labels)  # should be 14

pop.ids <- colnames(Vibrio.input.all.outlier.af)  # "F1.R1.freq", ..., "F63.R63.freq"

meta <- data.frame(
  sample_id   = pop.ids,
  bio_label   = Vibrio.labels,
  stringsAsFactors = FALSE
)

#Derive treatment from the biological label
meta$treatment <- ifelse(grepl("CC", meta$bio_label), "control_only",
                  ifelse(grepl("CS", meta$bio_label), "stress_only",
                  ifelse(grepl("VC", meta$bio_label), "vibrio_only",
                  ifelse(grepl("VS", meta$bio_label), "vibrio+stress", "initial"))))

#can also parse by probe set
meta$probes <- ifelse(grepl("Combined", meta$bio_label), "CADO+Vibrio",
                  ifelse(grepl("Vibrio", meta$bio_label), "Vibrio", ""))

View(meta)

#check that column names in input.all.rand.af and sample_id column in meta table are the same
all.equal(colnames(Vibrio.input.all.outlier.af), meta$sample_id)
# should be TRUE

# assign the columns of input.all.outlier.af with their proper sample names instead of the sample_id created by inputting the table to poolseq
colnames(Vibrio.input.all.outlier.af) <- meta$bio_label
```

```{r}
# Perform PCA
af_t <- t(Vibrio.input.all.outlier.af)

# Run PCA on transposed allele frequency matrix (samples as rows, loci as columns)
pca_res <- prcomp(af_t, center = TRUE, scale. = TRUE)

# Get scores for plotting
pca_df <- as.data.frame(pca_res$x)
#pca_df$locus <- rownames(input.all.outlier.af)  # keep locus IDs if you have them

# Calculate percent variance explained by each PC
var_explained <- pca_res$sdev^2 / sum(pca_res$sdev^2) * 100

pca_df$Treatment <- meta$treatment
pca_df$Probes <- meta$probes

outpca_vib <- ggplot(pca_df, aes(x = PC1, y = PC2, color = Treatment, fill=Treatment)) +
  geom_point(aes(alpha=0.2), size =7,shape=21, col="black")+  
  scale_fill_manual(values=treatPalette,name="Treatment")+ 
guides(alpha=FALSE) +theme(axis.title.x = element_text(size = 24),axis.title.y = element_text(size = 24)) +
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 1)
    ) +
  labs(
    x = paste0("PC1 (", round(var_explained[1], 1), "%)"),
    y = paste0("PC2 (", round(var_explained[2], 1), "%)"),
    title = "PCA of Outlier Loci (Vibrio)"
  ); outpca_vib

ggsave(filename = "outlier_PCA/outlier_PCA_Vibrioprobe.png", plot = outpca_vib, width = 6, height = 5, units = "in", dpi = 300)
```


# Manhattan Plots

## Per population Outlier Manhattan Plot Composite
```{r}
library(dplyr)
library(ggplot2)
library(patchwork)

make_manhattan_plot <- function(pv_data, sig_data, con_filter, allcon, pop_name,
                                outlier_color, y_limit = NULL) {
  
  pv_data <- pv_data %>%
    mutate(
      Group = case_when(
        SNP %in% con_filter$SNP ~ "Control Treatment Filtered",
        SNP %in% allcon$SNP ~ "Control Treatment Filtered",
        TRUE ~ "Neutral"
      )
    )
  
  man <- ggman(
    pv_data,
    pvalue = "QVC",  #update for treatment!
    chr = "CHR",
    snp = "SNP",
    bp = "BP",
    relative.positions = TRUE,
    sigLine = NA,
    pointSize = 1.5,
    title = "",
    alpha = 0.5
  )
  
  dfm <- man[[1]]
  dfm$chrom_alt <- factor(dfm$chrom_alt, levels = c(0, 1))
  
  dfmo <- semi_join(dfm, sig_data, by = "SNP")
  dfmo$Group <- "Outlier"
  
  dfmsplit <- split(dfm, dfm$chrom)
  xbreaks <- sapply(dfmsplit, function(x) {
    midpoint <- length(x$index) / 2
    if (midpoint < 1) midpoint <- 1
    x$index[midpoint]
  })
  
  y_max <- if (is.null(y_limit)) max(dfm$marker + 1, na.rm = TRUE) else y_limit
  
  p <- ggplot(dfm, aes(x = index, y = marker, colour = as.factor(chrom_alt))) +
    geom_point(aes(alpha = marker, shape = Group, size = marker)) +
    geom_point(
      data = dfmo,
      aes(x = index, y = marker, shape = Group, size = marker),
      color = outlier_color,
      fill = outlier_color
    ) +
    scale_x_continuous(
      breaks = xbreaks,
      labels = names(xbreaks),
      expand = c(0, 0),
      limits = c(0, max(dfm$index) + 10)
    ) +
    guides(colour = FALSE, alpha = FALSE, fill = FALSE, size = FALSE) +
    scale_y_continuous(expand = c(0, 0), limits = c(0, y_max)) +
    scale_size_continuous(range = c(0.25, 3)) +
    scale_alpha_continuous(range = c(0.05, 0.8), limits = c(0, 50)) +
    ggtitle(pop_name) +
    labs(x = "Chromosome", y = expression(-Log10(q-value))) +
    theme_classic() +
    scale_shape_manual(values = c(15, 21, 16, 17)) +
    scale_color_manual(values = c("grey", "dark grey"), guide = FALSE) +
    guides(shape = guide_legend(override.aes = list(color = "grey", size = 5, linetype = 0)))
  
  list(plot = p, dfm = dfm, dfmo = dfmo)
}

```

```{r}
vibrio_con_filter <- read.table("ToFilter.Con.Loci.Vibrio.txt", header = TRUE)

#load in control filtered 1% FDR loci for each population
total.sig.Vibrio.filt <- read.table("Total.Significant.Loci.Vibrio", header = TRUE)

```

> nrow(total.sig.ARC)
[1] 276
> nrow(total.sig.ARC.filt)
[1] 268 control filtered outliers (p<0.01)
> nrow(total.sig.NEH)
[1] 2216
> nrow(total.sig.NEH.filt)
[1] 1321 control filtered outliers (p<0.01)
> nrow(total.sig.MV)
[1] 4180
> nrow(total.sig.MV.filt)
[1] 2627 control filtered outliers (p<0.01)


```{r}
vib.h <- subset(total.sig.Vibrio.filt, Sig.VC == TRUE)


vib_res <- make_manhattan_plot(
  pv_data = Vib.pv,
  sig_data = vib.h,
  con_filter = vibrio_con_filter,
  allcon = allcon.Vibrio,
  pop_name = "Vibrio",
  outlier_color = "#3F8F7A"
)

vib_res

```


```{r}
png(filename = "ManhattanPlots/All_Populations_Manhattan_newfilt_2.png", 
    units = "in", width = 10, height = 6,  res = 300, bg = "transparent")
arc_res$plot / neh_res$plot / mv_res$plot +
  plot_annotation(tag_levels = "A") +
  plot_layout(guides = "collect") 
dev.off()
```



## ARC - Outlier Manhattan Plot
```{r}
arc_con_filter <- read.table("ToFilter.Con.Loci.ARC.txt", header = TRUE)
```

```{r, warning =FALSE, eval = FALSE}
arc.h <- subset(total.sig.ARC, Sig.STR == TRUE)
arc.c <- subset(ARC.pv, QCON < 0.1)


ARC.pp <- ARC.pv %>%
  mutate(
    Group = case_when(
      #SNP %in% singleton$SNP ~ "Singleton",
      SNP %in% arc.c$SNP ~ "Control Treatment Filtered",
      SNP %in% allcon.ARC$SNP ~ "Control Treatment Filtered",
      SNP %in% arc_con_filter$SNP ~ "Control Treatment Filtered",
      TRUE ~ "Shared SNP"
    )
  )

man <-ggman(ARC.pp, pvalue="QSTR", chr="CHR", snp="SNP",bp="BP", relative.positions = TRUE , sigLine = NA,
            pointSize=1.5,title = "", alpha = 0.5)

dfm <- man[[1]]

dfm$chrom_alt <- factor(dfm$chrom_alt, levels=c(0,1))

dfmo <- semi_join(dfm, arc.h, by = "SNP")
dfmo$Group <- "Outlier"

dfmsplit <- split(dfm, dfm$chrom)
xbreaks <- sapply(dfmsplit,function(x) {
    midpoint <- length(x$index)/2
    if(midpoint <1) midpoint <- 1
    return(x$index[midpoint])
})
arc.STR <-ggplot(dfm, aes(x= index, y=marker, colour = as.factor(chrom_alt)))+
  geom_point(aes(alpha=marker, shape=Group, size = marker))+  
  geom_point(data=dfmo, fill=cbPaletteSmall3[1], aes(x=index, y=marker, shape=Group, size=marker),  color = cbPaletteSmall3[1])+
  scale_x_continuous(breaks = xbreaks, labels = names(xbreaks),expand = c(0,0),limits = c(0,max(dfm$index)+10)) +
  guides(colour = FALSE,alpha=FALSE, fill = FALSE, size = FALSE) +
  scale_y_continuous(expand = c(0,0),limits=c(0,max(dfm$marker+1,na.rm=TRUE)))+
  scale_size_continuous(range = c(0.25,3)) +
  scale_alpha_continuous(range=c(0.05,0.8), limits= c(0,50))+
  ggtitle("ARC") +
  labs(x = "Chromosome") +
  theme_classic()+
  labs(y=expression(-Log10(q-value))) +
  scale_shape_manual(values=c(15,21,16,17))+
  scale_color_manual(values=c("grey", "dark grey"), guide =FALSE)+
  guides(shape = guide_legend(override.aes = list(color = "grey", size = 5, linetype =0) ) )

print(arc.STR)

ggsave(
  filename = "ARC_manhattan.png",
  plot     = arc.STR,
  width    = 10,      # in inches
  height   = 5,
  dpi      = 300,
  bg       = "transparent"  # or "white"
)
```






# Outlier Fst tables
## Pairwise Fst

```{r}
library(poolfstat)
```

Construct outlier sync files for each population!
```{bash, eval=FALSE}
#pull CHROM & BP from outlier loci file and sorts them into position file
# keeps header from sync file and prints only rows whose "CHROM"_"POS" matches the positions file 
awk 'BEGIN{FS=OFS="\t"} NR>1{print $2,$3}' Total.Significant.Loci.ARC | sort -u > ARC.outlier.positions.txt && awk 'BEGIN{FS=OFS="\t"} NR==FNR{if(FNR>1) keep[$1"_"$2]=1; next} FNR==1 || ($1"_"$2 in keep)' ARC.outlier.positions.txt ARC.dp25.cov.sync > ARC.outliers.only.sync

#check that these match the correct # of outliers
wc -l ARC.outliers.only.sync 

#remove extra lines with coverage stats
awk 'BEGIN{FS=OFS="\t"} {NF-=3; print}' ARC.outliers.only.sync > ARC.outliers.only.nocov.sync

# remove header line 
mawk '!/CHR/' ARC.outliers.only.nocov.sync > ARC.total.outlier.input
```


## ARC
```{r}
##Using sync file with outlier for each population 
out_ARC_samples <- c("ARC2_CB1","ARC2_CB2","ARC2_CB3","ARC2_SB1","ARC2_SB2","ARC2_SB3","ARC2_T01","ARC2_T02","ARC2_T03")

out_ARCdstbn <- popsync2pooldata("ARC.total.outlier.input", poolsizes = c(40000, 40000, 40000, 40000,
                                                                          40000, 40000, 40000, 40000, 40000), 
                                 poolnames = out_ARC_samples, min.rc = 1, min.cov.per.pool = 1, 
                                 max.cov.per.pool = 50000, min.maf = 0.1, noindel = TRUE, nlines.per.readblock = 1000)

#Compute Fst for every SNP
out_ARCdstbn_SNPFst <- computeFST(out_ARCdstbn, method = "Anova")

#Compute pairwise Fst matrix
out_ARCdstbn_Fst <- compute.pairwiseFST(out_ARCdstbn, method = "Anova", min.cov.per.pool = 1, max.cov.per.pool = 50000, min.maf = 0.001, output.snp.values = FALSE)

out_ARCdstbn_PairwiseFSTMatrix <- as.matrix(as.data.frame(out_ARCdstbn_Fst@PairwiseFSTmatrix))

ARC_Fst <- out_ARCdstbn_PairwiseFSTMatrix[c("ARC2_CB1","ARC2_CB2","ARC2_CB3","ARC2_SB1","ARC2_SB2","ARC2_SB3","ARC2_T01","ARC2_T02","ARC2_T03"), c("ARC2_CB1","ARC2_CB2","ARC2_CB3","ARC2_SB1","ARC2_SB2","ARC2_SB3","ARC2_T01","ARC2_T02","ARC2_T03")]
write.csv(ARC_Fst, "ARC_Fst_Outliers.csv")
```









# Neutral loci
### Prep Neutral loci files
Make neutral loci sync file
```{bash, eval=FALSE}
awk 'BEGIN{FS=OFS="\t"} { NF-=3; print }' VibrioProbed.dp25.cov.sync > VibrioProbed.dp25.cov.noStats.sync
awk 'NR==FNR { if (FNR>1) bad[$1]=1; next }
     FNR==1 { print; next }
    { key=$1"_"$2; if (!(key in bad)) print }' \
Total.Significant.Loci.Vibrio VibrioProbed.dp25.cov.noStats.sync > Vibrio.neutral.only.sync

```

```{bash}
wc -l Vibrio.neutral.only.sync
```

## Neutral Loci Fst
```{bash}
#remove extra lines with coverage stats
#awk 'BEGIN{FS=OFS="\t"} {NF-=3; print}' ARC.neutral.only.sync > ARC.neutral.only.nocov.sync

# remove header line 
mawk '!/CHR/' Vibrio.neutral.only.sync > Vibrio.total.neutral.input
```

```{r}
##Using sync file with neutral for each population 
neut_ARC_samples <- c("ARC2_CB1","ARC2_CB2","ARC2_CB3","ARC2_SB1","ARC2_SB2","ARC2_SB3","ARC2_T01","ARC2_T02","ARC2_T03")

neut_ARCdstbn <- popsync2pooldata("ARC.total.neutral.input", poolsizes = c(40000, 40000, 40000, 40000,
                                                                          40000, 40000, 40000, 40000, 40000), 
                                 poolnames = neut_ARC_samples, min.rc = 1, min.cov.per.pool = 1, 
                                 max.cov.per.pool = 50000, min.maf = 0.1, noindel = TRUE, nlines.per.readblock = 1000)

#Compute Fst for every SNP
neut_ARCdstbn_SNPFst <- computeFST(neut_ARCdstbn, method = "Anova")

#Compute pairwise Fst matrix
neut_ARCdstbn_Fst <- compute.pairwiseFST(neut_ARCdstbn, method = "Anova", min.cov.per.pool = 1, max.cov.per.pool = 50000, min.maf = 0.001, output.snp.values = FALSE)

neut_ARCdstbn_PairwiseFSTMatrix <- as.matrix(as.data.frame(neut_ARCdstbn_Fst@PairwiseFSTmatrix))

ARC_Fst <- neut_ARCdstbn_PairwiseFSTMatrix[c("ARC2_CB1","ARC2_CB2","ARC2_CB3","ARC2_SB1","ARC2_SB2","ARC2_SB3","ARC2_T01","ARC2_T02","ARC2_T03"), c("ARC2_CB1","ARC2_CB2","ARC2_CB3","ARC2_SB1","ARC2_SB2","ARC2_SB3","ARC2_T01","ARC2_T02","ARC2_T03")]
write.csv(ARC_Fst, "ARC_Fst_Neutral.csv")
```



## Neutral Loci PCA
```{bash, eval = FALSE}
# Activate the CASE conda environment
source activate CADO

mawk '!/CHR/' Vibrio.neutral.only.sync | cut -f1,2 | sort | uniq > neutral.pca.filter.Vibrio

mawk 'NR==FNR{a[$1,$2]; next} ($1,$2) in a' neutral.pca.filter.Vibrio VibrioProbed.dp25.cov.sync | cut --complement -f18- > Vibrio.neutral.pca.sync
```

```{r}
# Read sync file of significant neutral loci for neutral PCA
ARC.input.all.neutral <- read.sync(file="Vibrio.neutral.pca.sync", gen=seq(1,14), repl=seq(1,14), polarization="rising")
ARC.input.all.neutral.af <- af(ARC.input.all.neutral,gen=seq(1,14), repl=seq(1,14))

Vibrio.labels <- c(
  "CC1_TVEVibrio",   "CC2_TVEVibrio",   "CC3_TVEVibrio",   "CON_TV0Vibrio", 
  "CS1_TVEVibrio",   "CS2_TVEVibrio",  "CS3_TVEVibrio",   "STR_TV0Vibrio",
  "VC1_TVEVibrio",   "VC2_TVEVibrio",   "VC3_TVEVibrio",   "VS1_TVEVibrio",
  "VS2_TVEVibrio",   "VS3_TVEVibrio"
)
length(Vibrio.labels)  # should be 14

pop.ids <- colnames(ARC.input.all.neutral.af)  # "F1.R1.freq", ..., "F63.R63.freq"

meta <- data.frame(
  sample_id   = pop.ids,
  bio_label   = Vibrio.labels,
  stringsAsFactors = FALSE
)

#Derive treatment from the biological label
meta$treatment <- ifelse(grepl("CC", meta$bio_label), "control_only",
                  ifelse(grepl("CS", meta$bio_label), "stress_only",
                  ifelse(grepl("VC", meta$bio_label), "vibrio_only",
                  ifelse(grepl("VS", meta$bio_label), "vibrio+stress", "initial"))))

#can also parse by probe set
meta$probes <- ifelse(grepl("Combined", meta$bio_label), "CADO+Vibrio",
                  ifelse(grepl("Vibrio", meta$bio_label), "Vibrio", ""))

View(meta)

#check that column names in input.all.rand.af and sample_id column in meta table are the same
all.equal(colnames(ARC.input.all.neutral.af), meta$sample_id)
# should be TRUE

# assign the columns of input.all.neutral.af with their proper sample names instead of the sample_id created by inputting the table to poolseq
colnames(ARC.input.all.neutral.af) <- meta$bio_label
```

```{r}
# Perform PCA
af_t <- t(ARC.input.all.neutral.af)

# Run PCA on transposed allele frequency matrix (samples as rows, loci as columns)
pca_res <- prcomp(af_t, center = TRUE, scale. = TRUE)

# Get scores for plotting
pca_df <- as.data.frame(pca_res$x)
#pca_df$locus <- rownames(input.all.neutral.af)  # keep locus IDs if you have them

# Calculate percent variance explained by each PC
var_explained <- pca_res$sdev^2 / sum(pca_res$sdev^2) * 100

pca_df$Treatment <- meta$treatment
pca_df$Trial <- meta$trial

neutpca_arc <- ggplot(pca_df, aes(x = PC1, y = PC2, color = Treatment, fill=Treatment)) +
  geom_point(aes(alpha=0.2), size =7,shape=21, col="black")+  
  scale_fill_manual(values=treatPalette,name="Treatment")+ 
guides(alpha=FALSE) +theme(axis.title.x = element_text(size = 24),axis.title.y = element_text(size = 24)) +
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 1)
    ) +
  labs(
    x = paste0("PC1 (", round(var_explained[1], 1), "%)"),
    y = paste0("PC2 (", round(var_explained[2], 1), "%)"),
    title = "PCA of Neutral Loci (ARC)"
  );neutpca_arc

ggsave(filename = "neutral_PCA/neutral_PCA_Vibrioprobe.png", plot = neutpca_arc, width = 6, height = 5, units = "in", dpi = 300)
```
